---
title: "If I Only Had A Brain: Memory Retention After Decapitation and Regeneration in Planaria"
blank-lines-above-title: 2
shorttitle: "Memory Retention in Planaria"
author: 
  - name: Francis Forde
    email: francis.forde@vuw.ac.nz
    affiliations:
      - name: Victoria University of Wellington
        department: School of Psychology
        address: PO Box 600, Easterfield Building, Kelburn Pde, Kelburn
        city: Wellington
        region: New Zealand
author-note:
  blank-lines-above-author-note: 1
  disclosures:
    study-registration: ~
    # Acknowledge and cite data/materials to be shared.
    data-sharing: ~
    # Example: This article is based on data published in [Reference].
    # Example: This article is based on the dissertation completed by [citation].  
    related-report: ~
    # Example: [Author name] has been a paid consultant for Corporation X, which funded this study.
    conflict-of-interest: The authors have no conflicts of interest to disclose.
    # Example: This study was supported by Grant [Grant Number] from [Funding Source].
    financial-support: ~
    # Example: The authors are grateful to [Person] for [Reason].
    gratitude: ~
    # Example. Because the authors are equal contributors, order of authorship was determined by a fair coin toss.
    authorship-agreements: ~
abstract: "XX"
# Put as many keywords at you like, separated by commmas (e.g., [reliability, validity, generalizability])
keywords: [Memory, Learning, Planaria, Reinstatement, Cocaine, Behaviour]
# If true, tables and figures are mingled with the text instead of listed at the end of the document.
floatsintext: true

#Use to set position of footnotes
reference-location: margin

# Numbered lines (.pdf and .docx only)
numbered-lines: false
# File with references
bibliography: bibliography.bib
bibliographystyle: apa
# Suppress title page
suppress-title-page: false
# Link citations to references
link-citations: true
# Masks references that appear in the masked-citations list
mask: false
masked-citations: ~
# Language options. See https://quarto.org/docs/authoring/language.html
lang: en
language:
  citation-last-author-separator: "and"
  citation-masked-author: "Masked Citation"
  citation-masked-date: "n.d."
  citation-masked-title: "Masked Title"
  email: "Email"
  title-block-author-note: "Author Note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
  references-meta-analysis: "References marked with an asterisk indicate studies included in the meta-analysis."

format:
  apaquarto-docx: default
  apaquarto-html: 
    css: styles.css
  apaquarto-typst: 
    keep-typ: true
  
#format:
  #html:
    #css: styles.css
  #apaquarto-docx: default
  #apaquarto-typst: 
    #keep-typ: true
  apaquarto-pdf:
    # Can be jou (journal), man (manuscript), stu (student), or doc (document)
    documentmode: man
    fontsize: "12pt"
    # a4 paper if true, letter size if false
    a4paper: false
    # suppresses loading of the lmodern font package
    nolmodern: false
    # Suppresses loading of the fontenc package
    nofontenc: false
    # Suppress the title above the introduction
    donotrepeattitle: false
    # In jou mode, use times or pslatex instead of txfonts
    notxfonts: false
    # In jou mode, use Computer Modern font instead of times
    notimes: false
    # In jou mode, cancels automatic stretching of tables to column width 
    notab: false
    # Uses Helvetica font in stu and man mode
    helv: false
    # In man and stu mode, neutralizes the \helvetica command
    nosf: false
    # In man and stu mode, uses typewriter font
    tt: false
    # Puts draft watermark on first page
    draftfirst: false
    # Puts draft watermark on all pages
    draftall: false
    # Masks references that are marked as the author's own
    mask: false
    journal: ~
    volume: ~
    course: ~
editor: visual
fig-dpi: 300
fig-format: png
toc: true
toc-location: "body"
execute: 
  echo: false
  warning: false
  error: false
---

```{r}
#| label: global configuration and setup
#| include: false

#push updates using git push origin master

#To pull updates git checkout origin/master -- Francis_Masters_Thesis.qmd

# to push updates git commit -m "updating thesis manuscript"


#removes scientific notation (e.g. 10 x e^2)
options(scipen = 999)

# Conflicted
if (!require("conflicted")) install.packages("conflicted")
library(conflicted)

if (!require("pwr")) install.packages("pwr")
library(pwr)

# Tidyverse
if (!require("tidyverse")) install.packages("tidyverse")
library(tidyverse)

# Flextable
if (!require("flextable")) install.packages("flextable")
library(flextable)

# ftExtra
if (!require("ftExtra")) install.packages("ftExtra")
library(ftExtra)

# Knitr
if (!require("knitr")) install.packages("knitr")
library(knitr)

# Here
if (!require("here")) install.packages("here")
library(here)

# Patchwork
if (!require("patchwork")) install.packages("patchwork")
library(patchwork)

# Rempsyc
if (!require("rempsyc")) install.packages("rempsyc")
library(rempsyc)

# PNG
if (!require("png")) install.packages("png")
library(png)

# Grid
if (!require("grid")) install.packages("grid")
library(grid)

# JPEG
if (!require("jpeg")) install.packages("jpeg")
library(jpeg)

# ggplot2
if (!require("ggplot2")) install.packages("ggplot2")
library(ggplot2)

# APA
if (!require("apa")) install.packages("apa")
library(apa)

# Ghibli
if (!require("ghibli")) install.packages("ghibli")
library(ghibli)

# ggdist
if (!require("ggdist")) install.packages("ggdist")
library(ggdist)

# RColorBrewer
if (!require("RColorBrewer")) install.packages("RColorBrewer")
library(RColorBrewer)

# dplyr
if (!require("dplyr")) install.packages("dplyr")
library(dplyr)

# ggpubr
if (!require("ggpubr")) install.packages("ggpubr")
library(ggpubr)

# tidyr
if (!require("tidyr")) install.packages("tidyr")
library(tidyr)

# ggridges
if (!require("ggridges")) install.packages("ggridges")
library(ggridges)

# viridis
if (!require("viridis")) install.packages("viridis")
library(viridis)

# agridat
if (!require("agridat")) install.packages("agridat")
library(agridat)

# Conflicted
if (!require("conflicted")) install.packages("conflicted")
library(conflicted)

# showtext
if (!require("showtext")) install.packages("showtext")
library(showtext)



#library(conflicted)
#library(tidyverse)
#library(flextable)
#library(ftExtra)
#library(knitr)
#library(here)
#library(patchwork)
#library(rempsyc)
#library(png)
#library(grid)
#library(jpeg)
#library(ggplot2)
#library(apa)
#library(ghibli)
#library(ggdist)
#library(RColorBrewer)
#library(dplyr)
#library(ggpubr)
#library(tidyr)
#library(ggridges)
#library(viridis)
#library(agridat)
set_here() #sets the working directory



conflicts_prefer(dplyr::filter, .quiet = TRUE)
conflicts_prefer(flextable::separate_header, .quiet = TRUE)
conflicts_prefer(lme4::lmer, .quiet = TRUE)

```

# Introduction

-   Brief intrduction to area

-   Knowldge gap

-   How I addressed this

-   What I found

-   What this might mean

# Background

A brain in isolation is just a clump of extravagant cells. A brain earns its keep by liasing with the body and the external world. It is among these brain-environment interactions that an organism can set and achieve goals and, ultimately, carve a pathway to survival. But brains operate in the dark. Their only insight into the on-goings of the world is through delicately placed sensory organs such as the eyes, nose and ears. The sensory technology that each organisms possess', what philosophers call its sensorium, differs across species. Some build a picture of the world by capturing light using light sensitive proteins. Others live where no light can penetrate and so must form their worldview using other sensory modalities like echolocation. Notwithstanding these differences, neuroscience (in collaboration with biology) seeks to understand the suite of abilities each organism possesses, the neuronal and molecular mechanisms which underpin these, and the factors that determine when and why an organism deploys the behaviours in its arsenal. From the perspective of science, When we use non-human organisms we often do so while hoping the knowledge gleaned will teach us something about our own brains and bodies.

We now have at hand a broad tool set for inspecting the brain across different time spans and at different levels of analysis. From looking at activity within a single dendritic spine across microseconds to looking at connectivity between different brain structures over many seconds. We can even track changes in the size and number of individual spines on a single dendrite over time – impressive given the width of a spine is 100 times smaller than the thickness of a human hair [@bayramoglu2022; @li_current_2023]. At the network level, we are able to identify groups of neurons (ensembles) involved in encoding and storing memory, and can use precise tools to excite or inhibit those networks to alter an animals behaviour [@goshen_optogenetic_2014].

Our current experimental competency arose from many small steps. Before we had the means to manipulate neurons to identofy their role in memory, we had to make do with simple procedures like learning lists of nonsense syllables or simple motor tasks. This research often utilised patients who had conditions like amnesia or who had suffered some kind of brain injury.This early research helped answer the question of whether memory is a unitary system or a suite of separate systems which can be dissociated. Distinctions such as episodic and procedural, as well as short- and long-term memory emerged. Theoretical progress provided the foundation upon which specialised tools and procedures could be developed to manipulate and characterise the biology of memory in its different forms.

## Overview of key concepts in the field of learning and memory

### The meaning of memory and the different forms it takes

Generally speaking, memory is the embodiment of past experience which shapes our future behaviour. Learning can be thought of as the process of memory acquisition. Though, it must be said that there may be as many different definitions of learning and memory as there are papers published on the topic. @barron_embracing_2015 surveyed the various uses of the term "learning" across disciplines such as cognitive psychology, behavioural ecology, and machine learning and identified at least 50 definitions (albeit with a lot of overlap). Memory has been parceled into several distinct categories based on the content of the information held [see @fig-figure1 below]. A major distinction was made between explicit (or declarative) and implicit (or procedural) memory [@squire_memory_1987; @schacter_memory_1994]. Explicit memories are those accessible to consciousness. Implicit memories refer to information that you cannot consciously access but which still affect behaviour. Explicit memory has been further subdivided into episodic and semantic memory [@tulving_episodic_1972]. With episodic referring to the rich experiential quality of personal memories, while semantic relates to things that you know, but that lack an experiential component (e.g. facts about the world).

Memories can also be categorised on the temporal dimension. @spence_human_1968 proposed three stores which capture memory states over time: a sensory register, a short term, and a long term store. This division is still usefully applied in the field of learning and memory [e.g, @miller_timescales_2024]. A temporal framing of memory reflects the process of learning itself. Learning is normally broken up into several stages, each building on prior stages to further ingrain the memory and increase its longevity (see @sec-mechanisms below) . Initially, information enters a short term state and may alter behaviour and decision making in the immediate future. However, most of the information that our brains initially perceive – all the light hitting our eyes, and all the vibrations in the air registered by our ears – is not retained. To successfully retain this information it must be actively incorporated and maintained so that it can be accessed or recalled in perpetuity. To put it another way, the Information that is taken on board but must undergo specific processes (known as consolidation) to stave off the forgetting process.

These distinctions make a lot of sense when talking about memory in humans. But it may not be immediately clear what relevance these distinctions have when considering other organisms. Most people do not attribute rich episodic (experiential) memories to rodents, let alone invertebrates. Yet, it is clear that even very simple organisms have the capacity to learn and store varied types of memory; Ranging from basic changes in reflexes (habituation and sensitisation) to more complex operant behaviours. Such distinctions may be particularity relevant for the phenomena of memory storage outside of the brain, which is the topic pursued in this project. If, as recent research suggests [@shomrat_automated_2013], some memories are able to be retained outside the brain (and perhaps outside the CNS) these categories of memory may help us derive the parameters for which memory types hold this property and which do not. It may be that procedural memories such as non-associative responses can be stored outside of neural networks. Whereas memories which are consciously accessible can only be stored among complex ensembles of neurons in the brain. We do not yet know the bounds memory storage outside of the central nervous system. Armed with these conceptual distinctions of memory, we can investigate a broad range of training techniques to find out what forms of learning persist outside the brain and which do not.

```{r, eval = TRUE}
#| label: fig-figure1
#| fig-cap: "Categorisation of Memory"
#| apa-note: "Theroertical categoriasation of memory based on the content and cosncious accessibility of memory. The major explicit/implicit distinction was first put forward by Endel Tulving (1972). Figure based on Squire (1987)"
#| out-width: "100%"
#| fig-align: "center"

library(png)
library(grid)
Figure1 <- readPNG("Figures/Figure1.png") 
grid.raster(Figure1)
```

### Associative and non-associative learning

Associative learning requires learning the relationship between two stimuli. For example, one stimulus reliably precedes another, or a behaviour reliably precedes a reward. Non-associative forms of learning captures learning about a stimulus itself, but not in relation to other stimuli. This typically takes the form of behavioural sensitization or habituation. If you were to mildly shock my hand, I would withdraw it reflexively as it would be startling. But with repeated administration of the shock over time, I may come to learn that the shock is actually not that painful after all. The size of my startle response decreased over time with repeated exposure (habituation). I have learned something about the shock, but have learned nothing about its relationship to other stimuli. To extend this example to an associative form of learning, the shock could be delivered after being shown a picture of a certain flower. I would therefore associate (likely subconsciously) the flower imagery with a negative experience, and would eventually display a fear startle to presentations of the flower (tense up, squint my eyes, dip my head). This captures the fact that I have learned a temporal association between the flower and the shock. My body learns to prepare for the shock before it arrives.

Classical conditioning and operant conditioning are two frequently used forms of associative learning. Classical conditioning involves learning an association between two or more stimuli, as in the flower/shock example above. Operant conditioning differs from classical conditioning in that rather than a stimulus (e.g. shock) being paired with another stimulus (flower), a behaviour comes to be associated with a specific outcome. For example, I learn that signing up to psychology experiments often involves being exposed to (mildly) irritating stimuli. This changes the likelihood of that response being produced in the future. In other words, I stop signing up as a participant. Good for me, not so good for postgraduate students trying to complete their research.

### Maladaptive learning

The examples outline above capture learning in cases where it is beneficial for the learner. Preparing for shocks in the case of classical conditioning reduces the painfulness of the stimuli. Avoiding future experiments would mean you are exposed to less unnecessary irritants. These are simple examples, but we will all encounter consequential cases whereby our wellbeing and longevity is greatly enhanced by learning from past experiences. A close call when crossing a road (and the negative physiology experience that ensues) may increase the likelihood of diligently looking both ways in the future before crossing a road. These learning experiences allow us to persist in a world full of novel and ancient dangers. But this important capacity for learning raises several vulnerabilities.

Learning has a darker side. Consider Rebecca's first experience with heroine. Prior to consumption, Rebecca has heard about heroine, but only in the sense that she knows it is a harmful drug, that similar drugs are used in a medical setting, and so on. She has no prior subjective experience of its effects. But after several experiences with the drug, she learns about the intense sense of euphoria that comes from taking the drug. Later on, Sarah starts to feel a strong motivation to take the drug again, especially when she sees drug associated cues (syringes, white powder etc.). Without the capacity to form such associations, addiction would not be an issue. Sarah would fail to remember what actions led to such euphoric experiences, and no motivations would be aroused at the sight of drug paraphernalia. Addiction arises from the often useful ability to remember what past actions and events resulted in positive and negative experiences. In Rebecca's case, the euphoria that was experienced as positive during initial experiences led to neurological rewiring, hijacking the cognitive system to pursue heroine in the face of adverse consequences in the future. This is an unfortunate but not uncommon case of maladaptive learning.

Many aspects of poor mental health conditions, including aspects of depression, PTSD, and anxiety, similarly rely on memory systems for their development. Anxiety is especially relevant, as it involves the worry or concern over some perceived threat, often in response to environmental cues. The threat has not yet occurred, but to a person with anxiety, memory of past experience of similar circumstances creates excessive fear and a sense of unease as they imagine a similar negative outcome occurring in the future. By understanding what leads to the creation of these associations between a context and a negative outcome, memory research can then help us explore ways to prevent anxiety-like responses in the future, and identify methods to unlearn the initial negative association. But to arrive at insights that can be translated into practical steps for treating humans we must first start with simple tractable model organisms.

### Mechanisms of memory storage {#sec-mechanisms}

As described earlier, the process of acquiring an enduring memory involves transforming the information from a temporary state to an enduring state. The default process for all organisms is to forget information. In fact, we probably forget most of the information we encounter. This is evolutionary sensible. Storing information takes energy and physical realestate, and attaining energy comes at a price. Every organism therefore has a small number of things about which it must study extensively and track over time. But most of the information it encounters, be it visual, tactile, or olfactory, does not matter to it. A penguin cares about the scent of a polar bear, but not about the small bug being crushed under its foot. While a penguin's brain may process information about both, it is not equally likely to store both experiences and the contexts in which they took place.

When discussing how information is stored, one becomes steeped in complex biological pathways. Such pathways involve proteins interacting with other proteins, proteins interacting with DNA, and the production of new proteins. In the neurobiological literature, much of the discussion around learning takes place at the level of synapses – synapses being the points where two neurons interact. The axon from neuron A attaches to a dendrite from neuron B and thus form a synapse. Synapses are the locus of neuronal communication in the brain. At an abstract level, learning is realised by information affecting neurons such that they progress through a sequence of stages: generation, stabalisation, consolidation, and maintenance [see @rudy_neurobiology_2014 for a digestible overview of each stage]. There are complicated mechanisms involved in each stage of learning, and not all the components are fully understood. Despite that, the literature provides a good sense of many important molecular events that are thought to be involved in the process of storing information in the brain so that it can be accessed in the future.

When some new bit of information has been processed by the brain, the physical embodiment in the brain is often described as the "memory trace" [@asok_molecular_2019; @robins_21st_2023; @semon_mneme_1921]. An influx of calcium (resulting from stimulation by an upstream neuron A) is necessary for the initial **generation** of the memory trace [@lynch_intracellular_1983; @cavazzini_ca2_2005]. The calcium interacts within the post-synaptic dendritic spine on neuron B and leads to actin being disasembled into smaller chunks [@lynch_ltp_2007]. Actin helps give structure to the spine. Breaking it down is a necessary step to create room to fit more receptors into the post-synaptic membrane of neuron B – the membrane acts as the skin between the cell and the watery world outside of the cell. Adding receptors to the membrane makes the spine on neuron B more sensitive to the upstream firing of neuron A, and thus more likely to itself fire an action potential in response. But this rapid change in sensitivity (potentiation) is short lived. Without additional actions, this potentiation will revert back to baseline. **Stabilisation** of the trace requires expanding and strengthening the actin network in the dendritic spine to solidify the modifications [@chen_changes_2007]. The spine head is enlarged, and the actin scaffolding is changes to make it less vulnerable to being broken down (depolymerated). Furthermore, proteins in the cell membrane called adhesion molecules help to couple the pre- and post-synaptic neurons, improving the effectiveness of neurotransmission [@huntley_cadherin_2002]. These first two stages, generation and stabalisation, are like the pencil marks of memory. But these scratchings must be committed to ink to stand the test of time.

**Consolidation**, the stage where initial changes are solidified to stave off forgetting, involves the production of new proteins. In response to neuronal stimulation, a number of events take place in the posynaptic neuron. One response is the production of proteins which are able to enter the nucleus. Once in the neculeus, these proteins bind to DNA. In association with other molecular signalling, this activity leads to transcription of new molecules (mRNA) that will then be turned into proteins, such as receptors that can be inserted in the membrane. This genomic signalling ensures new proteins are continually minted, providing a sufficient pool of receptors and other elements needed to keep the spine in a potentiated (more sensitive) state. During the subsequent **maintenance** stage this supply of membrane receptors is produced and inserted into the membrane. and the typical dynamics of receptor removal, where receptors are removed from the membrane and recycled, is slowed. More excitatory receptors remain on the cell membrane which ensures heightened sensitivity to signals from presynaptic neurons both now and in the future.

### Memory research in animals and invertebrates

Many organisms have being poked, prodded and leered at in our attempts to understand the mechanisms underlying memory. Scrub jays, a bird sporting a bold blue coat and pointed black beak, have long been subjects in studies investigating spatial memory because of their food caching expertise [@shettleworth_how_1982]. Rather sophisticated methods have been used to study changes in hippocampal volume with caching and the possibility of season-dependent changes in hippocampal neurogenesis in caching birds [reviewed in @pravosudov_25the_2007]. Rodents have also featured heavily in the experimental memory literature [@ghafarimoghadam_review_2022]. Recent advances in stimulation and imaging, specifically techniques like optogenetics [@goshen_optogenetic_2014] and two-photon microscopy [@kawakami_vivo_2015], have enabled us to study representation of different types of memory at levels ranging from individual synapses to neuronal ensembles.

Rodent models have also been developed for tapping into spatial memory, working memory, associative memory, and object recognition. The last two decades saw a growing interest in episodic memory in rodents. Episodic memory, at its essence, is the ability to represent the past and draw on specific encoded events in a manner akin to mental time travel [@tulving_episodic_2002; @eacott_mental_2007; @krause_episodic_2022]. Intricate tasks have been developed which enable rats to demonstrate memory for the context in which a stimulus had been previously presented, and to disentangled this from mere familiarity with the stimulus due to temporal proximity [@panoz-brown_rats_2016]. However, the existence of episodic-like memory in non-human animals remains controversial [@tulving_episodic_2005; @hoerl_thinking_2019]. The establishment of procedures for identifying and manipulating complex episodic memory, by optogenetic and other means, may help inform us on the type of mechanisms (synaptic or molecular) that underpin episodic memory in humans.

Research in birds and rodents has supplied us with decades of insights into behaviour and the brain structures involved in memory. But when seeking to understand the precise molecular and structural changes that underpin memory, the field first looked to invertebrates. The simplicity of invertebrate neural architecture allowed researchers to account for and track the entirety of a nervous system, and to observe the functional specificity of individual neurons. But simplicity is not the only benefit. By using simpler organisms costs can be significantly reduced and environmental variables can be more easily controlled. Moreover, ethical concerns are relieved (albeit not completely removed) due to the reduced likelihood of finding meaningful sentience at this level.

While perhaps largely unknown outside of the sciences, aplysia is a celebrity among the invertebrates for its contribution to the neurobiology of learning. Aplysia is a marine snail with a simple nervous system. The abdominal ganglion (collection of neurons) of the aplysia is home to the largest known neurons in nature [@moroz_aplysia_2011]. This makes it an ideal candidate for electrophysiology – the approach where neural activity is recorded by inserting electrodes into cells or in the space surrounding cells. In aplysia, stimulation of the siphon (used for transporting water in the body) leads to a defensive retraction of the gill [@carew_classical_1981]. Repeated stimulation leads to a decrease in the intensity and length of the response, a simple form of non-associative memory called habituation[@pinsker_habituation_1970]. This simple form of learning is not something we typically care about. Yet, even this basic for of learning in aplysia served as a useful platform for understanding the stages of learning from generation through to consolidation described above, which likely translate to and underpin more complex forms of memory that we do care about in humans.

C. elegans is another organism which towers above most invertebrates in its popularity. C. elegans gained prestige after it was the first organisms to have its connectome mapped [@white_structure_1986]. Understating the wiring of all 302 neurons in C. elegans allowed for a a systems perspective of the nervous system. We could piece togeather how each neuron contributes to the overall whole and enables the body to perform a suite of actions like movement, digestion, and defensive behaviours. Though the scale C. elegans nervous system is small compared to mammals, it revealed principles of neuronal organisation which persist across brains of all sizes. Principles such as reciprocal inhibition to facilitate movement, computing at the level of the cell for efficiency, and minimising the total length of neuronal wire [@sterling_principles_2015]. Although we may find comfort in distancing ourselves from so-called "lower organisms", nature is indifferent to our need for preeminence. Nature seeks to wire up machines that survive. Effective blueprints are often reused.

Our study of invertebrates revealed that complex behaviour can arise from a modicum of neural cells. Consider that C. elegans has less connections in its entire nervous system (\~7000) than a single mammalian pyramidal neuron has synaptic connections [@cook_whole-animal_2019; @sterling_principles_2015; @megias_total_2001]. Yet, this bare-bones neuronal setup is sufficient for detecting a variety of chemical and olfactory cues, navigating an environment, escaping threats and detecting dynamic environmental signals such as temperature and social crowding. And as we climb the ladder of complexity from C. elegans to more complex invertebrates, the cognitive capabilities and potential for translational insights expands.

### Planaria as a model organism

Planaria are a broad group of invertebrates which have become the centerpiece for a number of scientific investigations. Planariara are being used to investigate topics including regenerative biology [@karami_planarians_2015], toxicology [@li_effects_2008; @hagstrom_comparative_2019], radioprotective materials [@ermakov_planarians_2021] addiction [@raffa_planaria_2008], and the effect of zero gravity environments on morphology [@vista_ssep_mission_11_team_studying_2018]. Planaria have built niches across many ecological contexts and can be found across salt-water, fresh-water and land environments. Land dwelling planaria span up to half a meter long [@esser_land_1981], whereas freshwater planaria, which are more commonly used in behavioural research, are less than a centimeter in length [@vasquez-doorman_current_2022].

Planaria are bilaterians. They display bilateral symmetry across their left and right sides [@sluys_planarian_2018]. Planaria exhibit anterior-posterior polarity, such that their head can be distinguished from the tail in both structure and their behavioural repertoire. While the tail end of a planarian is rather uninteresting, the head has many intriguing features. Auricles, the sensory organs capable of chemoreception, are what give the head of many planarian species its triangular shape. These provide planaria with their ability to detect food and noxious chemicals in the immediate environment. Eyespots, which are the most discernible feature of planaria, sit atop the dorsal surface of the head. These light-sensitive cell clusters enable planaria to detect light intensity and its direction[@shettigar_hierarchies_nodate]. Of particular interest to neruoscientists, the planarian head harbours a bilobed brain which is needed to coordinate activity throughout the body[@inoue_planarian_2015]. This simple neural structure is of special evolutionary significance as planaria are thought to be the oldest organism to house an organised central nervous system, or what we might label as a true brain [@pagan_first_2014; @sarnat_brain_1985]. In real terms, a planarian brain lacks many features compared to the exuberance of the mammalian brain. But in relative terms, the brain-to-body mass ratio of planaria is similar to that of a rat [@best_transphyletic_1983].

The planarian brain resembles a horseshoe[@agata_structure_1998; @sarnat_brain_1985] and has been estimated to contain between twenty to thirty thousand neurons [@inoue_functional_2017, p.82]. The brain exhibits nine branches on each side which radiate out from the center. The lobes of the brain form thin nerve cords at their posterior end. These chords extend down the length of the body towards the tail, and together with the brain, comprise the central nervous system. The left and right nerve cords are connected by commissures which form a ladder-like structure [@sluys_planarian_2018]. At first glance, the nerve chords do not have the organisational complexity of the planarian brain. Judging purely from appearances, the nerve chords should not be capable of performing complex behaviour. However, It has been shown that even cells in the nerve chords can take on the behavioural profile of a head, indicating that the nerve chords are not simply axons innervating muscle [@le_planarian_2021].

Planarian neurons appear more similar in structure to those of vertebrates than of other invertebrates [@sarnat_brain_1985]. They feature spine-like protrusions on dendrites [@sarnat_brain_1985; @petralia_diversity_2016], and many neurons contain many dendritic branches but a singular axon. Zooming in further, planarian neurons contain a variety of synaptic vesicles, such as clear and dense-core variations, which resemble those seen in vertebrate neurons [@oosaki_observations_1965]. Most relevant to the research described in this project, planaria produce many of the same neurotransmitters and neuromodulators that we humans posses. These include serotonin, dopamine, epinephrine, acetylcholine, GABA, glutamate and opioid peptides [@sarnat_brain_1985; @welsh_monoamine-containing_1970; @rawls_measurement_2006; for a comprehensive review of planarian neurochemistry see @buttarelli_neuropharmacology_2008].

The physiologist August Krogh posited that "You will find in the lower animals mechanisms and adaptations of exquisite beauty and the most surprising character" [-@krogh_progress_1929, p.203]. The conservation of neurochemistry in planaria is intriguing. But the primary feature making planaria worthy of Krogh's dictum is their regenerative capability. Planaria undergo a natural form of fission as part of their reproductive cycle. They tear themselves in half, with each half then regrowing all the necessary parts of its basic body plan to form a complete planarian again. A form of reproductive immortality. Regeneration is not completely novel in nature. Humans can regrow skin, and salamanders can regrow amputated limbs. But what sets planaria apart from the rest of the natural world is its ability to regrow its brain and central nervous system. This regeneration is facilitated by adult pluripotent neoblast cells which are found throughout the body [@neuhof_vertically-_2016]. After significant injury, these cells proliferate and undergo differentiation, providing the cell types needed to restore organs, membranes, and neural networks in the brain. This capability has drawn interest from medical researchers for more than a century [@morgan_experimental_1898; @child_patterns_1941; @reddien_cellular_2018], who realised that by understanding the factors that control planarian regeneration, we may be able to artificially simulate these processes in humans to restore limbs or neural structures after injury. We have now reached the stage where business ventures like the company Morphoceuticals are being established to apply lessons learnt from planarian regeneration to rodents and, pending pre-clinical success, eventually humans.

### Review of planarian memory literature

In the early 1900's, a common belief was that invertebrates did not have the cognitive means needed to learn. People conceded that perhaps local tissue change may allow some acute form of memory. But this would inevitably be erased or overwritten due to subsequent changes. Invertebrates were thus thought to lack permanent memory systems. This view is evident in the strong statement provide by Donald Jensen in the 1970s that "no invertebrate, no matter how complex is capable of showing 'true learning'" [quoted in @rilling_mystery_1996, p. 591]. This view established an artificial barrier separating the types of animals that are suitable models of human cognition from those that are not. Because invertebrates were off the table, researchers attempted to make progress on the neurobiology of memory using dizzyingly complex animals such as rodents. After many years searching for the rodent engram (the collection of neurons underlying a specific learning event), this venture unearthed little if anything of value. A group of psychologists including James McConnell in the 1970s were aware that little progress was being made in this endeavour. The group then moved away from rodents and, looking past the anthropocentric restrictions imposed on cognition, drifted towards invertebrates. Starting with much simpler organisms would allow more insightful discoveries, as small neural systems could be mapped and understood as a first step, while complicated mammalian nervous systems could not. The situation was akin to assembling a rocket ship for exploring space before figuring out commercial aviation within the earths atmosphere.

At first, McConnell and colleagues completed basic experiments showing that planaria could learn to associate a light with a shock. Compared to control subjects, trained planaria would exhibit more body contractions in response to light (a conditioned stimulus) and also perform more changes of direction. But criticism arose over the lack of controls in these experiments. Later follow ups included blinding the experimenter and testing for confounding factors such as pseudoconditioning (the eliciting of the unconditioned response by other stimuli despite there being no relationship between them) and sensitisation (an increase in responding to a CS due to repeated presentation, rather than because of its association with a US). Contrary to the expectations of psychologists at the time, evidence for learning in invertebrates accrued study after study. It was eventually impossible to deny the capacity for stable associative memories to these (admittedly) primitive looking creatures. McConnell and others such as Eric Kandel established definitively that invertebrates are capable of learning, retaining, and acting on information.

The ability to form associative memories is an impressive feat given the bare-bones neural networks of the planarian brain. But learning research could be carried out in thousands of other organisms. It was the pairing of the capacity to learn with the rare ability for regeneration that sprouted one of the most peculiar branches of scientific endeavor to date - investigating the retention of memory after decapitation and regeneration of the brain. Because planaria could learn simple associations, researchers wee able to ask the question: What happens if you teach a planarian, then cut it in half? Does the tail, which needs to regenerate its head and central nervous system, retain any of that prior learning? James McConnell was one of the first people to pose and pursue the answer to this question. Across a range of different training procedures, McConnell found that planarian tails indeed retain information, suggesting the intuition that memories must be stored in the brain may be mistaken, at least in some cases. Through some mechanism, memories are stored or backed up outside the brain and can be reinstantiated in the new brain during regeneration.

After the McConnell chronicles of the early 60s, invertebrate memory research saw its modern resurgence in the work of Shomrat and Levin [-@shomrat_automated_2013]. The authors published an important paper which utilised an automated training protocol for planaria. Planaria, like rodents, are hesitant to approach food in the center of a novel environment. They will first explore the territory, and only then engage in consumption. As planaria become familiar with the environment through repeated trials, they begin to approach the food more quickly, demonstrating a form of recognition memory. The authors wanted to know whether this type of memory persists in the tails of trained planaria after they have been cut in half and had to regenerate a brain from scratch.

Planaria can detect and distinguish surface textures – an ability referred to as thigmotaxis. Shomrat and Levin's study therefore used smooth and rough surface textures for the conditioned stimuli. Over ten consecutive days, half of the planaria were fed on the target rough surface ("familiar" planaria) while the other half were only fed on a smooth surface ("naive" planaria). At the end of the training period, the familiar group took a significantly shorter amount of time to approach and consume the food in the rough environment. Both groups were then bisected into head and tail halves ("offspring") and left to regenerate for 10-14 days. The authors then looked at whether the tail offspring of familiar planaria retained familiarity of the rough environment and thus approached food relatively quickly compared to the naive tail offspring. The data suggested that regenerated fragments from familiar planaria did approach food more quickly, but this did not reach statistical significance. Whats more, after undergoing the same training procedure as the original planaria, the regenerated tail fragments from familiar planaria demonstrated a form of memory savings. The familiar tail sections became accustomed to the rough environment faster than tail sections of control (unfamiliar) planaria. This hints that some memory trace from prior training survived brain regeneration, but required repetition of the training process for the memory savings to be expressed.

More recently, Samuel and colleagues [-@samuel_addiction-related_2021] corroborated this puzzling memory retention effect. The authors used sucrose to shift the surface preference of some planaria from their innate preference for a smooth surface to the sucrose-paired rough surface. After amputating the planaria and allowing time for head regeneration, it was observed that the tail halves retained the sucrose-paired rough preference, despite the newly regenerated brain never having been exposed to the rough surface. In contrast, the tail halves of control planaria – which were exposed to the rough surface but did not receive sucrose in this environment – showed the expected initial preference for the smooth surface.

One assumption implicit in planaria experiments to date is that, while a brain is not needed for storage of all memories, it is necessary to act upon the memories. For this reason, sufficient time is given for the brain to regenerate, enabling the planarian to swim and feed. However, a recent preprint by Shimojo and colleagues [-@shimojo_preservation_2022] goes against this assumption. They sought to test whether dissected tail halves of planaria can display a conditioned response (which was learnt before dissection) before the bran has a chance to regenerate. In this study, planaria were trained to associate a neutral weak UV light (conditioned stimulus) with an aversive shock (unconditioned stimulus). The shock typically causes planaria to twist their body – an unconditioned contortion response. After pairing the light with the shock, planaria will display a conditioned contortion response to the UV light alone. On the second and third day after dissection, well before the brain is thought to be reformed, the tail halves were exposed to the conditioned stimulus over a number of trials and their responses were recorded. The authors analysed the data using a deep neural network to classify behaviour. They found that most responses from the tail halves were similar to those produced by an electric shock rather than those produced by a neutral ultraviolet light. Ultimately, this suggests the tail halves retained the conditioned behaviour and were able to act on it, despite lacking a brain at the time.

Following a similar procedure, Rhodes and Vierick [-@rhodes_effects_2024] used classical conditioning to establish conditioned negative phototaxis (moving away from the conditioned light stimulus). Typically, planaria are strongly averse to blue light, mildly averse to green light, and show no natural aversion (or attraction) to red light [@paskin_planarian_2014]. Planaria were trained to associate a neutral red light with an aversive green light across 5 days. After conditioning, some planaria were bisected into head and tail halves. Three weeks later, the head and tails were tested for retention of the memory – negative phototaxis to red light. The results indicate that the bisected halves retained the conditioned memory just as well as intact planaria. Moreover, memory retention was not statistically different for regenerated heads and tails. This suggests planaria can retain a conditioned response for at least three weeks. What's more, the tail halves which lost their head and therefore had to regenerate a head and brain also retained and acted on this memory. One major issue with this study was that the number of planaria per group was very small, most containing just four to six subjects. Another key issue is it was not clear how much movement was necessary to support negative phototaxis on a given trial.

Although classical conditioning procedures are common in the planarian literature, some experimenters have employed operant conditioning methods [@crawford_operant_1967; @chicas-mosier_new_2015; see @best_behavior_1963 for a review of early studies]. A simple learning procedure known as the Van Oye was one of the first forms of reinforcement learning in planaia [@van_oye_over_1920; @wells_training_1967; @raffa_analysis_2008]. In the typical setup, planaria are housed in a beaker and a fishing line with food attached is placed near the surface of the water. Planaria can detect the presence of food through chemo sensory receptors. Planaria must navigate up the wall, across the surface and down the line to reach the food. This is a low probability behaviour, but a small to medium percentage of planaria will eventually do this given enough trials. The rod is lowered gradually across training trials, so that planaria will learn to descend to lower depths on the line to reach the food. Control planaria undergo similar methods but without the food reward attached. At test, food is not placed on the rod, but is instead dissolved in the water beforehand. The dissolved food is a cue that food is available. Trained planaria are subsequently found in much greater numbers on the suspended line compared to control subjects. Across five experiments performed by Wells, an average of \~17 trained subjects were found on the line at test compared to an average of \~3 experimental subjects [reviewed in @corning_planarian_1970]. This procedure demonstrates that planaria can be trained using reinforcement learning.

Another operant conditioning study was conducted by Corning [-@corning_retention_1966] during the zenith of planaria fame. Corning wondered whether planaria can retain an operant conditioned behaviour after bisection and regeneration. Using a T-maze apparatus, planaria were trained to select their less preferred side. Being returned to their home cage for 10 minutes was used as a positive reinforcer for the correct choice. Otherwise, they were taken to the start of the maze for another trial. Corning decided on a threshold for successful learning of nine out of ten consecutive correct choices across trials. After being dissected and left to regenerate for two to three weeks, the offspring (heads and tails) were given a baseline preference test and were subsequently conditioned to criterion. Corning found that the baseline of trained offspring differed significantly from the baseline of the original planaria, while untrained planaria offspring did not differ from the original subjects. This suggested operant conditioned behaviour can be retained outside of the planarian brain. Extending this, the regenerates of trained planaria could also be conditioned to threshold faster than regenerates of untrained planaria. A demonstration of direct memory recall and memory savings.

### Cocaine as a positive reinforcer for Planaria

Investigators have used many different stimuli, both aversive and appetitive, in their behavioural studies. One of the most common appetitive stimuli in the planarian literature is cocaine. Cocaine is an ideal candidate drug for behavioural studies given it acts on dopamine receptors which are abundant in planaria [@algeri_effects_1983; @buttarelli_neuropharmacology_2008]. Moreover, cocaine can be easily dissolved in water at low cost given the small quantity needed to reward planaria. The literature shows that cocaine induces strong effects on locomotion and atypical behaviours when given at high doses [@rawls_first_2010; @pagan_planarians_2013].

Cocaine is a stimulant which exerts it agonistic effects by blocking reuptake of dopamine through the dopamine transporters. In humans, this results in more dopamine activity in the synapse and therefore more neural activity in downstream neurons, particularly in the meso-limbic pathway connecting the ventral tegmental area to the nucleus accumbens [@nestler_molecular_2019]. This drug effect is linked to the rush or high that users experience, and is common to most if not all drugs of abuse. Cocaine also acts on serotonergic and noradrenergic transmission by blocking their respective transporters. The noradrenergic effects are thought to stimulate the sympathetic nervous system by blocking reuptake of norepinephrine and an decreased sympathetic nerve discharge, resulting in effects such as increased blood pressure and heart rate [@jacobsen_effects_1997; @freye_pharmacology_2009; @nestler_molecular_2019].

Amaning-Kwarteng et al., [-@amaning-kwarteng_relapse_2017] explored the establishment and extinction of a drug reinforced preference, and the ability to reinstate this preference after extinction. They found that planaria can be conditioned using cocaine to shift their surface texture preference (e.g from smooth to rough), and that this preference can be extinguished (reverted back to the original preference) after repeated exposure without reinforcement. Subsequently, exposure to a bath of cocaine was enough to reinstate the conditioned preference when provided with both textures and given the choice of which surface it spends its time. Building on this research, Jawad, Hutchinson & Prados [-@mohammed_jawad_dissociation_2018] investigated addiction like learning and memory behaviour using sucrose. As is common in the planarian literature, the authors employed a conditioned place preference paradigm and measured conditioning, extinction, and tolerance. Conditioned place preference is commonly used to assess the motivational strength of drugs [@milton_persistence_2012]. A novel aspect of the work by Jawad and colleagues was the demonstration that appetitive learning in planarians requires dopaminergic activity, evidenced by a dopamine D1 antagonist blocking acquisition of a CPP but not interfering with tolerance.The use of planaria to understand addiction related phenomena such as tolerance and extinction has benefited from a recent resurgence. But its roots date back to at least the 1960's [@needleman_tolerance_1967].

Understanding the molecular changes and circuit dynamics underlying the establishment of addiction may illuminate a path towards chemical or behaviour therapies to change the brain and thus change behaviour. Currently, therapies focus on top down strategies. People are coached to recognise their thoughts and emotions related to drugs, and to manage them rather than simply act on them. However, if the chemistry and structural wiring of the brain change during the acquisition of an addiction, bottom up therapies may assist in reverting these maladaptive neuroadaptations [@chodkiewicz_conceptual_2023]. And although we may not be in a position to attempt extreme interventions in humans or other mammals, planaria enable us to pursue these questions and progress our understanding of how bottom up changes may reduce drug seeking behaviour.

### Unresolved questions

In the first half of the 20th century, there was widespread doubt regarding whether invertebrates can learn. But as we look back nearly a century later, we have gathered ample evidence that planaria can learn [@wells_training_1967; @samuel_addiction-related_2021; @amaning-kwarteng_relapse_2017]. In addition, planaria serve as an especially useful organism due to their regenerative capabilities. Combined with conditioning procedures, there is now evidence that memory can be successfully retained in a planarian after having its head removed and thus having to regenerate a brain [@shomrat_automated_2013]. It is an extraordinary finding that a simple preference or learned aversion can be maintained within the body despite complete devastation of the central nervous system. But even more incredible would be the persistence of complex behaviour.

Although a conditioned texture preference is a form of learning, it is very rudimentary and feels distant from the types of human memories that concern us in our day to day lives. Meanwhile, learning to perform some action to receive a reward, requiring both memory systems and their modulation by dopamine circuitry, is much closer to home for us humans. It is much closer to the forms of learning that we care about and value, and view as a sign of intelligence. If complex memories formed by operant conditioning can persist in spite of total brain loss, this may have profound implications for the way we view memory storage and retrieval in humans. With that said, to my knowledge there are no clear examples of retention of operant conditioned behaviour through decapitation and regeneration in planaria. This project therefore attempts to generalise the principle of memory retention through regeneration shown for classical conditioning to an operant conditioning procedure.

A second uncertainty surrounding the memory retention phenomena is the type of post-bisection exposure needed to reinstantiate a given memory. Shomrat and Levin [-@shomrat_automated_2013] observed that regenerated planaria tails from trained subjects did not initially differ in their performance compared to controls. However, it was clear that their performance on the task improved more rapidly than tail halves of control planaria. The memory seemed to lay dormant but could be reactivated to some extent after exposure to the relevant training condition – a for of retention called "memory savings". This type of memory retention is similar to the procedure of reinstatement in addiction research. After successfully training an animal to lever press for a reward such as cocaine, extinction is achieved by allowing the animal to repeatedly engage in the lever pressing behaviour without being rewarded. Eventually, the animal will stop performing the conditioned response when the lever is presented. However, if the animal is exposed to the reinforcer before being placed back in the operant chamber, the lever pressing behaviour will spontaneously return.

With respect to both phenomena, the memory is not accessible or is not acted upon, and requires exposure to the right stimulus to be reactivated. Extinction and reinstatement of drug seeking behaviour has been shown in planaria [@amaning-kwarteng_relapse_2017]. But what remains to be seen is whether reinstatement can be used to reactivate memories which are dormant after decapitation and regeneration. The phenomena of savings demonstrates some memory trace is retained in the brainless tail half. Perhaps this can be reactivated promptly by exposing the planarian to the reinforcing stimulus before assessing for memory retention – an inveterate reinstatement procedure.

# Materials and Method

### Colony maintenance

The species of planaria used in this research is not currently known. Due to restrictions on importing identified species such as *Schmidtea mediterranea* into New Zealand, local planaria were sourced from two local streams within Wellington, New Zealand (X stream and Y stream). Given the basic characteristics of the planaria (colour, head shape etc.) it is thought that there is a combination of Cura and Neppia species -- both of which are commonly found in New Zealand waterways. We intend to perform genomic analysis at a later date to confirm the species identity. Planaria were housed in a 50 liter glass aquarium with internal filtering and which contained a natural ecological environment (rocks,snails,algae etc.). The tank water was maintained with -\*describe water treatment protocol\*- (referred to as "planaria water" hereafter). The room containing the aquarium was maintained at 23°C. The planaria were fed between one and three times a week, with meals consisting of \[describe food source\]. The colony was maintained on a 12-hour light/dark cycle with lights on at 9:30am till 9:30pm.

### Materials

**Experiment 1** (dose-response analysis): plastic petri dishes with a diameter of 5.5cm were used to assess motility. Petri dishes contained a final solution of 8ml, made up of pure planaria water for control subjects and cocaine hydrochloride (Sigma Aldrich, United States??) mixed with planaria water for experimental subjects. Planaria locomotion was captured using an OPPO A17 smart phone and videos were imported into EthoVision (Noldus Information Technologies, Wageningen, the Netherlands) for motility tracking. Most subjects could not be tracked for some duration due to being in a shadow or obscured by the petri dish wall (\~10% of frames on average). To resolve this, the missing data were interpolated using the interpolation feature in EthoVision which uses a direct line from the subjects last location to the next observed location to determine the distance traveled.

**Experiment 2** (Y-maze conditioning): three custom Y-mazes were used. The mazes were identical, and were made from 80x80mm squares, with the compartments laser etched into the plastic. The mazes contain a 27mm long runway, and two arms each being 25mm long. All compartments were 6mm wide and all walls were 6.5mm high. At the intersection between the runway and the arms, there is a small divot on the floor of the maze. This allows a plug 7mm in diameter to be inserted to trap liquid in the arms and enable controlled treatment with a drug. The maze floor contained subtle lines as a result of the etching process. At the base of the runway there is a small externally powered white light (\~20 lux) which was fixed into the plastic. The light is a mild aversive stimulus intended to induce negative phototaxis and discourage the planarian from resting at the start of the runway.

### Handling

Planaria were handled using several techniques based on the circumstances. When removing planaria from their 12-well-plate, a filbert paintbrush was preferred. However, when moving planaria between petri dishes and the y-maze, a fine artist's paintbrush was preferred. In other cases, such as when planaria would sit in the middle of the y-maze divot, a plastic transfer pipette with the tip cut off was used. Planaria were handled very gently throughout their lifespan. Rough handling was suspected to have caused a very high mortality rate during pilot experiments. This was likely due to puncturing the planaria by pointing the bristles at the body on approach. For the studies reported here, a different technique was adopted whereby the side of the bristles was used to gently scoop the planaria up the side of the wall until they attached their cilia to the brush.

## Experiment 1

Experiment 1 tested for an appropriate dose of cocaine which is rewarding to planaria while not drastically altering their their locomotive behaviour. Cocaine has been frequently used as a rewarding compound to classically condition planaria and to investigate its toxicity [@amaning-kwarteng_relapse_2017; @raffa_description_2005; @tallarida_ethanol_2014; @hutchinson_persistent_2015; @palladini_pharmacological_1996]. Doses used for conditioning have ranged from 1μM [@hutchinson_persistent_2015] to 80μM [@raffa_cocaine_2005]. The planarian species used across conditioning experiments vary, including *Dugesia Tigrina*, *Dugesia gonocephala* and *Dugesia dorotocephala*.

Investigators have observed that some species are more amenable to conditioning procedures than others [@samuel_addiction-related_2021; @mueller_use_2002]. Differences in the behaviour and responses of planarian species have been observed in response to several types of stimuli [@cochet-escartin_scrunching_2015; @debold_differences_1965]. The species used throughout this project likely differs from those used elsewhere in the literature and may in fact be a species indigenous to New Zealand. For this reason, it is important to identify a suitable dose of cocaine which is rewarding but does not significantly alter motility.

60 planaria were used in this experiment and were spread across six conditions based on doses commonly used in the planarian conditioning literature. This included 0, 1, 5, 10, 20 and 100μM (*n* = 10 per condition). Subjects were run across twelve session. The first ten sessions each contained one subject from the 0 - 20μM conditions, whereas the last two session only contained subjects in the 100μM condition[^1]. Subjects were collected from the breeding tank on the day of data collection. Within each session, subjects were randomly allocated to their condition using a freely available [random number generator](https://stattrek.com/statistics/random-number-generator#table).

[^1]: initially only 0 - 20μM conditions were run. The 100μM condition was added after the initial data were analysed to ensure that the cocaine was having some effect on the planaria and was not inert.

Each dose-response session lasted 15 minutes. Prior to the first recording session of the day, the drug concentrations were achieved by mixing cocaine (dissolved in distilled water) with planarian water to reach a final solution of 8ml. Each solution was mixed and allowed to sit for several minutes to ensure diffusion of the drug. Planaria were assigned to each condition from a large pool prior to the session. A planarian was picked up and a random number sequence was followed to determine which condition it was assigned to. The recording began once all five subjects were in their respective dishes. After completing a single trial, the planaria were rehoused in a large tank and were not used for any subsequent experiments in this manuscript.

@fig-figure2 provides an overview of the apparatus setup. Five petri dishes were positioned on a white acrylic sheet. Recording sessions took place under red light, with the light positioned 36cm above the petri dishes. The dishes were aligned in a 2x3 grid, with a gap left in the top middle position which is where the light was focused to minimise shadows in the petri dishes; this was important for subsequent digital tracking. Each drug concentration was rotated across the 5 grid positions between trials to control for any effects of lighting angle.

```{r, eval = TRUE}
#| label: fig-figure2
#| fig-cap: "Dose response apparatus"
#| include: true
#| fig-width: 12
#| fig-height: 7
#| fig-align: "center"
#Figure2
Figure2 <- readJPEG("Figures/Figure2.jpg") 
grid.raster(Figure2)
```

### Results

```{r}
#| label: Dose Response Descriptives
#| include: False

library(papaja)
library(rstatix)
library(afex)


# Source the script
source("Descriptives.R")
source("Graphs.R")
source("PvalueFormattingFunction.R")

```

@fig-boxplot depicts the distance moved by planarians across the six conditions. Prior to performing any statistics, the assumptions of normality and homogeneity of variances were tested. Levene's test for homogeneity of variances suggests there were equal variances across conditions (F = `r format(levene_dose_response$F, digits = 3)`, `r format_pvalue(levene_dose_response$P)`). The Shapiro-Wilk test results indicate that the data were not normally distributed (W = `r format(shapiro_test_dose_response$statistic, digits = 3)`,`r format_pvalue(shapiro_test_dose_response$p.value, digits = 3)`). Due to violation of the assumptions of ANOVA, a Kuskal test was performed. The results indicated that the effect of `Condition` on `Distance` is statistically significant (χ2 (`r kruskal_test$parameter`) = `r format(kruskal_test$statistic, digits = 3)`, `r format_pvalue(kruskal_test$p.value, digits = 3)`). An exploratory post-hoc Dunn's test was carried out to determine whether the 100μM group differed from other groups. The results indicated that the 100μM group differs significantly from several other groups: control (`r format_pvalue(dunn_test$P[11])`), 5μM (= `r format_pvalue(dunn_test$P[7])`), 10μM (`r format_pvalue(dunn_test$P[1])`), and 20μM (`r format_pvalue(dunn_test$P[4])`).

```{r, eval = TRUE}
#| label: fig-boxplot
#| fig-cap: "Plot of planarian motility by condition"
#| fig-width: 12
#| fig-height: 7
#| fig-align: center
#| warning: false
#| message: false

source("Plots/rainplot.R")
print(rainplot)

```

The results in @fig-boxplot convey the variability of planarian behaviour. All conditions had at least one subject which moved less than 30cm over the 15 minute recording, and all groups had at least two subjects that moved more than 140cm. Experimenter observations indicate that when placed in the recording dish, some planaria would move initially, and then come to rest within a few minutes at a spot on the wall. They would remain here without meaningful movement for the remainder of the recording. Although the 1μM group did not differ significantly from the control group, there is a curious grouping of planaria in the 1μM condition below 60cm. Consistent with this, Hutchinson et al., [-@hutchinson_persistent_2015] observed a significant decrease in motility during exposure to 1μM of cocaine but not to 10μM when compared to a control group. No potential explanation was offered for this unusual dose-response effect.

The dose-response results suggest that any doses between 1μM and 20μM would be appropriate for conditioning, as they should not significantly alter planarian behaviour. The next relevant criteria is to choose a dose that is sufficiently rewarding. A range of cocaine doses have been used to successfully condition planaria in CPP paradigms. These procedures typically involve low doses such as 1μM [@hutchinson_persistent_2015], 5μM [@amaning-kwarteng_relapse_2017] or 10μM [@hutchinson_persistent_2015]. But in the case of CPP, exposure time per trial is relatively long, on the order of 15-20 minutes. In the operant conditioning paradigm proposed in Experiment 2, exposure time will be 3 minutes. To adjust for the smaller absorption window compared to CPP experiments, the higher dose of 20μM was chosen for future conditioning experiments.

One interesting observation is that planaria tend to move much more at the start of the session than towards the end, with a gradual decrease in the distance moved as each minute passes (see @fig-ridgeplot). An exploratory Welch's two sample t-test found a significant difference between the time traveled in the first minute (M = `r format(minute1_descriptives$mean, digits = 3)`, SD = `r format(minute1_descriptives$sd, digits = 3)`) compared to the 15th minute (M = `r format(minute15_descriptives$mean, digits = 3)`, SD = `r format(minute15_descriptives$sd, digits = 3)`), with subjects travelling significantly further during minute 15 (t(`r (time_binned_t_test_result$parameter)`) = `r format(time_binned_t_test_result$statistic, digits = 3)`, `r format_pvalue(time_binned_t_test_result$p.value)`). In future dose-response assessments, a shorter recording duration may be adequate to assess the effects of a compound on motility.

```{r}
#| label: fig-ridgeplot
#| include: True
#| fig-cap: "Plot of planarian motility across recording interval"
#| apa-note: "Distanced moved by planaria during each minute of the 15-minute recording interval. Each ridge plots the distance moved for the entire sample during that minute bucket. Black bars indicate the mean distance moved across all conditions. "
#| fig-width: 12
#| fig-height: 7
#| fig-align: "center"

source("Plots/ridgeplot.R")

print(ridgeplot)

```

## Experiment 2

Next we sought to determine whether planaria can learn and retain an operant conditioned response. Prior research has demonstrated the capacity for learning by way of classical conditioning. And there is evidence that classically conditioned memories can be retained after decapitation and regeneration of the brain. But the capacity for complex memories shaped by operant conditioning to persist despite losing most of the central nervous system has not been definitively shown. As a first step towards testing retention of operant conditioned memory, we needed to first establish the capacity for operant learning in this species of planaria. This experiment was preregistered prior to data collection and can be found online at [Open Science Framework](https://osf.io/tq7u4/?view_only=9c794dd942fb4a54b6a986c0a893fe46) and at [PsycArchives](https://www.psycharchives.org/en/item/d6109ed1-9aab-467b-b981-e009be95f308).

Sixty planaria were used (cocaine group, *n* = 30, vehicle group, *n* = 30). This experiment had four stages: baseline, conditioning, test, and priming. for baseline and conditioning two planaria were run concurrently in two separate Y-mazes. The maze was filled with 1.8ml of planaria water. The water was shaken so as to distribute evenly throughout the maze, with bubbles removed as needed. Six planaria were used per run, wherein they completed either 6 trials (baseline) or 4 trials per day (conditioning) with an intertrial interval of approximately 15 minutes. The six planaria were first moved into holding petri dishes, with the regular white room light on to encourage movement. At the start of a trial, two planaria were transferred to the middle of the runway using a paintbrush, and shaken loose into the water. Once the planaria had entered the runway, the timer was started. Planaria were given three minutes to enter one of the arms [^2]. Once a planarian had entered an arm, the plug was inserted to stop liquid moving between compartments. Each arm contained 0.5ml after the plug was inserted. A planarian was considered to have entered the arm when the plug could be safely inserted without touching the planarian.

[^2]: If the planarian had some part of their body in an arm, they would be given up to an extra minute to make their decision

For treatment subjects, if the active arm was selected, 43.5μL of cocaine in distilled water was pipetted near the center of the arm. If the inactive arm was selected, an identical volume of distilled water was pipetted near the center of the arm. After administration, the timer was restarted and three minutes were given for absorption. For control subjects, either arm resulted in plugging and then pipetting 43.5μL of distilled water into the arm. If a subject failed to enter an arm, the plug was inserted and 43.5μL of distilled water was pipetted near the subjects body while in the runway and then three minutes were given. The runway light was on throughout the duration of the trial. At the end of a trial, planaria were gently removed and placed back into their holding dish.

At test, six planaria were used per run. Three planaria were run concurrently in three separate Y-mazes. Planaria were given three minutes to make a decision. Once a decision was made, the plug was inserted and planaria were left for \~60 seconds before being moved back to the holding dish. No additional liquid was added to any compartment of the Y-maze during the test trials. The next group of three planaria would then begin their first test trial. The inter trial interval was approximately six minutes and thirty seconds. For the priming stage, the procedure was identical to the test stage, with the added component of the drug exposure before the first trial. At the start of a run, the planaria were placed in a 8ml solution of cocaine diluted in planaria water at a conectration of 20μM for 10 minutes[^3]. At the end of the exposure interval, planaria were moved individually into a Y-maze to begin their first trial. Planaria were only exposed to cocaine before the first priming trial, but not before subsequent trials.

[^3]: To optimise efficiency, the start time for each group of 3 was staggered so that subjects 1-3 had finished their first trial before the 10-minute period ended for subjects 4-6.

#### Results and discussion

```{r}
#| label: Experiment Two Results
#| include: False


# Source the scripts
source("ExperimentTwoAnalysis.R")
source("ExperimentTwoTimeAnalysis.R")


```

@fig-exp2decisions shows the average percentage of trials where subjects entered the active arm across four stages. A generalised linear mixed effects model with family set to binomial was fitted in R. Subject ID was set as a random effect, with condition, time point and the interaction term as a fixed effects. Pairwise comparisons with a Bonferroni correction were carried out using the emmeans package in R. Type III Wald chi-square tests were carried out to identify whether there was a main effect of condition, time, and their interaction. The results did not find a significant effect of condition (χ2 (`r (exp2_decisions_model_output$Df[2])`) = `r format(exp2_decisions_model_output$Chisq[2], digits = 3)`,`r format_pvalue(exp2_decisions_model_output$"Pr(>Chisq)"[2])`). The results indicate a significant effect of time (χ2 (`r (exp2_decisions_model_output$Df[3])`) = `r format(exp2_decisions_model_output$Chisq[3], digits = 3)`, `r format_pvalue(exp2_decisions_model_output$"Pr(>Chisq)"[3])`) and a significant time\*condition interaction (χ2 (`r (exp2_decisions_model_output$Df[4])`) = `r format(exp2_decisions_model_output$Chisq[4], digits = 3)`, `r format_pvalue(exp2_decisions_model_output$"Pr(>Chisq)"[4])`).

Post-hoc pairwise comparisons were carried out using estimated marginal means with Bonferroni corrections applied to account for multiple comparisons. Pairwise comparisons were carried out based on the model-estimated response probabilities rather than the raw proportions. Comparisons looked at within group differences in response probability across the four phases and also at between group differences at each stage.The results indicated that there were two within-group differences for the control subjects: endpoint differed significantly from baseline (`r format_pvalue(exp2_decision_comparisons_within_group$p.value[1])`), and test differed significantly from endpoint (`r format_pvalue(exp2_decision_comparisons_within_group$p.value[4])`). There were three within-group differences for the treatment subjects: endpoint differed significantly from baseline (`r format_pvalue(exp2_decision_comparisons_within_group$p.value[7])`), test differed significantly from baseline (`r format_pvalue(exp2_decision_comparisons_within_group$p.value[8])`), and reinstatement differed significantly from endpoint (`r format_pvalue(exp2_decision_comparisons_within_group$p.value[11])`). A significant difference was found in the preference score between treatment and control groups at test (`r format_pvalue(exp2_decision_comparisons_between_group$p.value[3])`). \newpage

```{r}
#| label: fig-exp2decisions
#| include: True
#| fig-cap: "Mean percentage of choices for the active arm over time between conditions"
#| apa-note: "Change in y-maze arm preference across four experimental phases.A) mean proportion of trials where the active arm was chosen for the control subjects. There was a significant difference between baseline and endpoint, as well as endpoint and test. B) mean proportion of trials where the active arm was chosen for the treatment subjects. There were significant differences between baseline and endpoint, baseline and test, and endpoint and reinstaement. C) combined plot showing between group differences in the mean proportion of trials where the active arm was chosen. There was a significant difference in active arm preference at test. * = p<.05; ** = p<.01; *** = p<.001." 
#| fig-width: 12
#| fig-height: 14
#| fig-align: "center" 
#| dpi: 300

source("ExperimentTwoAnalysis.R")

print(combined_panel)

```

\newpage

Looking first at the results [@fig-exp2decisions, B], subjects in the treatment group show evidence of successful conditioning. These subjects were more likely to choose the active arm at the end of conditioning (endpoint) compared to baseline. Moreover, this preference was maintained for two weeks as evident by the heightened preference at the test point. Despite the increase in preference persisting for two weeks, when tested the next day during the reinstatement procedure the preference had returned to a level similar to that of baseline.

The control group [@fig-exp2decisions, A] demonstrated a similar increase in preference for the active arm despite receiving just the vehicle (distilled water) for making this choice. However, it contrast to the treatment group, the change in preference returned to levels seen at baseline when tested two weeks after conditioning. This highlights natural variability in planaria behaviour over time and introduces healthy skepticism against the claim that treatment planaria were successfully conditioned.

The between groups comparison seen in [@fig-exp2decisions, C] shows a significant difference between groups found during the test trials two weeks after the conditioning procedure. It may be that the persisting increase in preference shown by the treatment group is evidence of true learning as opposed to the natural variability of behaviour seen in the control group. This could explain why the treatment group effect remained even after two weeks, while the control group effect diminished back to baseline. The reinforcement day results constrain this line of thinking. Although conditioned responses have been shown to extinguish relatively quickly in planaria, reinstatement procedures have also been shown to be effective. Therefore, if the conditioned behaviour was able to persist for several weeks, there is little reason to expect it would be extinguished so quickly but fail to show the expected effect of reinstatement. Overall, the results do show some promise for the capacity of planaria to learn this y-maze procedure. Future replications and refinements may be needed to more convincingly demonstrate a learned response.

This experiment also considered the response time of planaria during each trial. It was hypothesised that although planaria may not be able to learn the complex operant behaviour, they may demonstrate increased motivation from being aware that an award is available. This could be inferred from faster responding in the treatment group compared to the control group. The response time data across the four phases are shown in [@fig-decision-time].

```{r}
#| label: fig-decision-time
#| include: True
#| fig-cap: "Mean time taken to make a response across phases and between conditions"
#| apa-note: "XX" 
#| fig-width: 16
#| fig-height: 14
#| fig-align: "center" 
#| dpi: 300

source("ExperimentTwoTimeAnalysis.R")

print(grouped_time_horizontal_boxplot)
```

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::

## Graveyard

.Some common examples of invertebrates studied within psychology include flies, bees, worms, and octopi

Planarians are thought to have existed for approximately 300 million years (Vila-Farré & Rink, 2018).

This allowed fine-grained control of experimental variables to minimise variability across trials.
