---
title: "If I Only Had A Brain: Memory Retention After Decapitation and Regeneration in Planaria"
blank-lines-above-title: 2
shorttitle: "Memory Retention in Planaria"
author: 
  - name: Francis Forde
    email: francis.forde@vuw.ac.nz
    affiliations:
      - name: Victoria University of Wellington
        department: School of Psychology
        address: PO Box 600, Easterfield Building, Kelburn Pde, Kelburn
        city: Wellington
        region: New Zealand
author-note:
  blank-lines-above-author-note: 1
  disclosures:
    study-registration: ~
    # Acknowledge and cite data/materials to be shared.
    data-sharing: ~
    # Example: This article is based on data published in [Reference].
    # Example: This article is based on the dissertation completed by [citation].  
    related-report: ~
    # Example: [Author name] has been a paid consultant for Corporation X, which funded this study.
    conflict-of-interest: The authors have no conflicts of interest to disclose.
    # Example: This study was supported by Grant [Grant Number] from [Funding Source].
    financial-support: ~
    # Example: The authors are grateful to [Person] for [Reason].
    gratitude: ~
    # Example. Because the authors are equal contributors, order of authorship was determined by a fair coin toss.
    authorship-agreements: ~
abstract: "XX"
# Put as many keywords at you like, separated by commmas (e.g., [reliability, validity, generalizability])
keywords: [Memory, Learning, Planaria, Reinstatement, Cocaine, Behaviour]
# If true, tables and figures are mingled with the text instead of listed at the end of the document.
floatsintext: true

#Use to set position of footnotes
reference-location: margin

# Numbered lines (.pdf and .docx only)
numbered-lines: false
# File with references
number-sections: true
bibliography: bibliography.bib
bibliographystyle: apa
# Suppress title page
suppress-title-page: false
# Link citations to references
link-citations: true
# Masks references that appear in the masked-citations list
mask: false
masked-citations: ~
# Language options. See https://quarto.org/docs/authoring/language.html
lang: en
language:
  citation-last-author-separator: "and"
  citation-masked-author: "Masked Citation"
  citation-masked-date: "n.d."
  citation-masked-title: "Masked Title"
  email: "Email"
  title-block-author-note: "Author Note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
  references-meta-analysis: "References marked with an asterisk indicate studies included in the meta-analysis."

format:
  apaquarto-docx: default
  apaquarto-html: 
    css: styles.css
  apaquarto-typst: 
    keep-typ: true
  
#format:
  #html:
    #css: styles.css
  #apaquarto-docx: default
  #apaquarto-typst: 
    #keep-typ: true
  apaquarto-pdf:
    # Can be jou (journal), man (manuscript), stu (student), or doc (document)
    documentmode: man
    fontsize: "12pt"
    # a4 paper if true, letter size if false
    a4paper: false
    # suppresses loading of the lmodern font package
    nolmodern: false
    # Suppresses loading of the fontenc package
    nofontenc: false
    # Suppress the title above the introduction
    donotrepeattitle: false
    # In jou mode, use times or pslatex instead of txfonts
    notxfonts: false
    # In jou mode, use Computer Modern font instead of times
    notimes: false
    # In jou mode, cancels automatic stretching of tables to column width 
    notab: false
    # Uses Helvetica font in stu and man mode
    helv: false
    # In man and stu mode, neutralizes the \helvetica command
    nosf: false
    # In man and stu mode, uses typewriter font
    tt: false
    # Puts draft watermark on first page
    draftfirst: false
    # Puts draft watermark on all pages
    draftall: false
    # Masks references that are marked as the author's own
    mask: false
    journal: ~
    volume: ~
    course: ~
editor: visual
fig-dpi: 300
fig-format: png
toc: true
toc-location: "body"
execute: 
  echo: false
  warning: false
  error: false
---

```{r}
#| label: global configuration and setup
#| include: false

#push updates using git push origin master

#To pull updates git checkout origin/master -- Francis_Masters_Thesis.qmd

# to push updates git commit -m "updating thesis manuscript"


#removes scientific notation (e.g. 10 x e^2)
options(scipen = 999)


#YOu may need Rtools installed to enable lme4 and Matrix package to be installed
#otherwise errors arise.

# Conflicted
if (!require("conflicted")) install.packages("conflicted")
library(conflicted)

if (!require("pwr")) install.packages("pwr")
library(pwr)

# Tidyverse
if (!require("tidyverse")) install.packages("tidyverse")
library(tidyverse)

# Flextable
if (!require("flextable")) install.packages("flextable")
library(flextable)

# ftExtra
if (!require("ftExtra")) install.packages("ftExtra")
library(ftExtra)

# Knitr
if (!require("knitr")) install.packages("knitr")
library(knitr)

# Here
if (!require("here")) install.packages("here")
library(here)

# Patchwork
if (!require("patchwork")) install.packages("patchwork")
library(patchwork)

# Rempsyc
if (!require("rempsyc")) install.packages("rempsyc")
library(rempsyc)

# PNG
if (!require("png")) install.packages("png")
library(png)

# Grid
if (!require("grid")) install.packages("grid")
library(grid)

# JPEG
if (!require("jpeg")) install.packages("jpeg")
library(jpeg)

# ggplot2
if (!require("ggplot2")) install.packages("ggplot2")
library(ggplot2)

# APA
if (!require("apa")) install.packages("apa")
library(apa)

# Ghibli
if (!require("ghibli")) install.packages("ghibli")
library(ghibli)

# ggdist
if (!require("ggdist")) install.packages("ggdist")
library(ggdist)

# RColorBrewer
if (!require("RColorBrewer")) install.packages("RColorBrewer")
library(RColorBrewer)

# dplyr
if (!require("dplyr")) install.packages("dplyr")
library(dplyr)

# ggpubr
if (!require("ggpubr")) install.packages("ggpubr")
library(ggpubr)

# tidyr
if (!require("tidyr")) install.packages("tidyr")
library(tidyr)

# ggridges
if (!require("ggridges")) install.packages("ggridges")
library(ggridges)

# viridis
if (!require("viridis")) install.packages("viridis")
library(viridis)

# agridat
if (!require("agridat")) install.packages("agridat")
library(agridat)

# Conflicted
if (!require("conflicted")) install.packages("conflicted")
library(conflicted)

# showtext
if (!require("showtext")) install.packages("showtext")
library(showtext)

#svglite
if (!require("svglite")) install.packages("svglite")
library(svglite)

#rstatix
if(!require("rstatix")) install.packages("rstatic")
library(rstatix)

#dunn.test
if(!require("dunn.test")) install.packages("dunn.test")
library(dunn.test)

#psych
if(!require("psych")) install.packages("psych")
library(psych)

#psych
if(!require("hrbrthemes")) install.packages("hrbrthemes")
library(hrbrthemes)

#psych
if(!require("hrbrthemes")) install.packages("hrbrthemes")
library(hrbrthemes)

#effectsize
if(!require("effectsize")) install.packages("effectsize")
library(effectsize)

#gghalves
if(!require("gghalves")) install.packages("gghalves")
library(gghalves)

#ggtext
if(!require("ggtext")) install.packages("ggtext")
library(ggtext)

#emmeans
if(!require("emmeans")) install.packages("emmeans")
library(emmeans)

#DHARMa
if(!require("DHARMa")) install.packages("DHARMa")
library(DHARMa)

#car
if(!require("car")) install.packages("car")
library(car)

#lme4
if(!require("lme4")) install.packages("lme4", type = "binary")
library(lme4)

#ggsignif
if(!require("ggsignif")) install.packages("ggsignif")
library(ggsignif)

#readxl
if(!require("readxl")) install.packages("readxl")
library(readxl)

#Matrix
if(!require("Matrix")) install.packages("Matrix", type = "binary")
library(Matrix)

#colorspace
if(!require("colorspace")) install.packages("colorspace")
library(colorspace)

set_here() #sets the working directory



conflicts_prefer(dplyr::filter, .quiet = TRUE)
conflicts_prefer(flextable::separate_header, .quiet = TRUE)
conflicts_prefer(lme4::lmer, .quiet = TRUE)

```

# Introduction

-   Brief introduction to area

-   Knowledge gap

-   How I addressed this

-   What I found

-   What this might mean

# Background

A brain in isolation is just a clump of extravagant cells. A brain earns its keep by liaising with the body and the external world. It is among these brain-environment interactions that an organism can set and achieve goals and, ultimately, carve a pathway to survival. But brains operate in the dark. Their only insight into the on-goings of the world is through delicately placed sensory organs such as the eyes, nose and ears.

The sensory technology that each organism possess', what philosophers call its sensorium, differs across species. Some build a picture of the world by capturing light using light sensitive proteins. Others live where no light can penetrate and so must form their worldview using other sensory modalities like echolocation. Notwithstanding these differences, neuroscience (in collaboration with biology) seeks to understand the suite of abilities each organism possesses, the neuronal and molecular mechanisms which underpin these, and the factors that determine when and why an organism deploys the behaviours in its arsenal. But we do not usually do this for the sake of the organisms itself. Rather, we use non-human organisms with the hope of learning something about our own brains and bodies.

We have at hand a broad tool set for inspecting brains across different time spans and at different levels of analysis. From looking at activity within a single dendritic spine over microseconds to looking at connectivity between different brain structures over several minutes. We can even track changes in the size and number of individual spines on a single dendrite over time – impressive given the width of a spine is 100 times smaller than the thickness of a human hair [@bayramoglu_hair_2022; @li_current_2023]. At the network level, we are able to identify groups of neurons (ensembles) involved in encoding and storing memory, and can use precise tools to excite or inhibit those networks to alter an animals behaviour [@goshen_optogenetic_2014].

Our current experimental competency arose from many small steps. Before we had the means to manipulate synapses and identify their role in memory, we had to make do with simple procedures like learning lists of nonsense syllables or simple motor tasks. This research often utilised patients who had conditions like amnesia or who had suffered some kind of brain injury. This early research helped answer the question of whether memory is a unitary system or a suite of separate systems which can be dissociated. Distinctions such as episodic and procedural, as well as short- and long-term memory emerged. Theoretical progress provided the foundation upon which specialised tools and procedures could be developed to manipulate and characterise the biology of memory in its different forms.

## Overview of key concepts in the field of learning and memory

### Categories of memory

Generally speaking, memory is the embodiment of past experience which shapes our future behaviour. Learning, on the other hand, is the process of memory acquisition. That said, there may be as many different definitions of learning and memory as there are papers published on the topic. @barron_embracing_2015 surveyed the various uses of the term "learning" across disciplines such as cognitive psychology, behavioural ecology, and machine learning and identified at least 50 definitions (albeit with a lot of overlap). Memory has been parceled into several distinct categories based on the content of the information held (see @fig-figure1 below). A major distinction was made between explicit (or declarative) and implicit (or procedural) memory [@squire_memory_1987; @schacter_memory_1994]. Explicit memories are those accessible to consciousness. Implicit memories cannot be consciously accessed but still affect behaviour. Explicit memory has been further subdivided into episodic and semantic memory [@tulving_episodic_1972]. Episodic refers to the rich experiential quality of personal memories, while semantic relates to things that you know, but which lack an experiential component such as facts about the world.

Memory can also be categorised temporally. @spence_human_1968 proposed a model with three stores that process memories over time: a sensory register, short-term store, and long-term store. This division is still usefully applied in the field of learning and memory [e.g, @miller_timescales_2024]. A temporal framing of memory reflects the process of learning itself. Learning is normally broken up into several stages, each building on prior stages to fortify the memory and increase its longevity (see @sec-mechanisms below). Initially, information enters a short-term state and may alter behaviour and decision making in the immediate future. However, most sensory information is not retained. Only meaningful information is given permanent residence. For this to occur, a set of active processes are required to ensure the information is maintained so that it can be accessed in perpetuity. This stage, known as consolidation, fights against the otherwise inevitable process of forgetting.

These distinctions make a lot of sense when referring to human memory. But it may not be immediately clear what relevance these distinctions have when considering other organisms. Most people do not attribute rich experiential memories to rodents, let alone invertebrates. Yet, even very simple organisms such as planaria (see @sec-planaria-as-a-model-organism) have the capacity to retain and act on information from their environment. Viewing memory as a range of dissociable information stores is particularity relevant when investigating whether information can be stored outside of the central nervous system. If, as recent research suggests [@shomrat_automated_2013], memories are able to be stored outside the brain, the categories previously described allow us to elucidate which forms of information have this property and which do not.

While non-associative memories such as sensitisation may be stored outside of neural networks, perhaps rich episodic experiences can only be stored among complex ensembles of neurons in the brain. We do not yet know the bounds memory storage outside of the central nervous system. Armed with these conceptual distinctions of memory, researchers can investigate a broad range of training techniques to identify which forms of memory persist outside the brain and which do not.

```{r, eval = TRUE}
#| label: fig-figure1
#| fig-cap: "Categorisation of Memory"
#| apa-note: "Theroretical categorisation of memory based on the content and conscious accessibility of the information. The major explicit/implicit distinction was first put forward by Endel Tulving (1972). Figure adapted from Squire (1987)."
#| fig-width: 8
#| fig-height: 5
#| fig-align: "center"

library(png)
library(grid)
Figure1 <- readPNG("Figures/Figure1.png") 
grid.raster(Figure1)
```

### Associative and non-associative learning

Associative learning requires learning the relationship between two stimuli. For example, one stimulus reliably precedes another, or a behaviour reliably precedes a reward. Non-associative forms of learning captures learning about a stimulus itself, but not in relation to other stimuli. This typically takes the form of behavioural sensitization or habituation. If you were to mildly shock my hand, I would withdraw it reflexively as it would be startling. But with repeated administration of the shock over time, I may come to learn that the shock is actually not that painful after all. The size of my startle response decreased over time with repeated exposure (habituation). I have learned something about the shock but have learned nothing about its relationship to other stimuli. To extend this example to an associative form of learning, the shock could be delivered after being shown a picture of a certain flower. I would therefore associate (likely subconsciously) the flower imagery with a negative experience, and would eventually display a fear startle to presentations of the flower (tense up, squint my eyes, dip my head). This captures the fact that I have learned a temporal association between the flower and the shock. My body learns to prepare for the shock before it arrives.

Classical conditioning and operant conditioning are two frequently used forms of associative learning. Classical conditioning involves learning an association between two or more stimuli, as in the flower/shock example above. Operant conditioning differs from classical conditioning in that rather than a stimulus (e.g. shock) being paired with another stimulus (flower), a behaviour comes to be associated with a specific outcome. For example, I learn that signing up to psychology experiments often involves being exposed to (mildly) irritating stimuli. This changes the likelihood of that response being produced in the future. In other words, I stop signing up as a participant. Good for me, not so good for postgraduate students trying to complete their research.

### Maladaptive learning

The examples outline above capture learning in cases where it is beneficial for the learner. In the classical conditioning case described above, preparing for a shock by tensing muscle tissue reduces the painfulness of the experience and minimises the chance of tissue damage. For the operant conditioning case, avoiding future experiments would mean you are exposed to less unnecessary irritants. Although these are mundane examples, we will all encounter consequential cases whereby our wellbeing and longevity is enhanced by learning from past experiences. A close call when crossing a road (and the negative physiology experience that ensues) may increase the likelihood of diligently looking both ways before future crossings. These learning experiences allow us to persist in a world full of both ancient and novel dangers. But learning is not always protective.

The capacity for learning leaves us vulnerable to developing maladaptive habits. Consider Rebecca's first experience with heroine. Prior to consumption, Rebecca has heard about heroine, but only in the sense that she knows it is a harmful drug, that similar drugs are used in a medical setting, and so on. She has no prior subjective experience of its effects. But after several experiences with the drug, she learns about the intense sense of euphoria that comes from taking the drug. Later on, Sarah starts to feel a strong motivation to take the drug again, especially when she sees drug associated cues (syringes, white powder etc.).

Without the capacity to form such associations, addiction would not be an issue. Sarah would fail to remember what actions led to such euphoric experiences, and no motivations would be aroused at the sight of drug paraphernalia. Addiction arises from the often useful ability to remember what past actions and events resulted in positive and negative experiences. In Rebecca's case, the euphoria that was experienced as positive during initial experiences led to neurological rewiring, hijacking the cognitive system to pursue heroine in the face of adverse consequences in the future. This is an unfortunate case of maladaptive learning.

Many aspects of poor mental health conditions, including aspects of depression, PTSD, and anxiety, can only develop because of our ability to learn. For example, anxiety entails worry or concern over some perceived threat. The threat has not yet occurred, but to a person with anxiety, past experience of similar circumstances creates excessive fear as they imagine a similar negative outcome occurring again. By understanding what biological mechanisms create associations between a context and a negative outcome, as in the case of anxiety, we can then move to develop methods which erode the maladaptive associations and reduce human suffering. Although we have made progress in our understanding of how memories are instantiated in biological beings, there are many unknowns which constrain our ability to create practical interventions.

## Mechanisms of memory storage {#sec-mechanisms}

As described earlier, the process of acquiring an enduring memory involves transforming the information from a temporary state to an enduring state. The default process for all organisms is to forget information. In fact, we probably forget most of the information we encounter. This is evolutionary sensible. Storing information takes energy and physical real estate, and attaining energy comes at a price. Every organism therefore has a small number of things about which it must study extensively and track over time. But most of the information it encounters, be it visual, tactile, or olfactory, does not matter to it. A penguin cares about the scent of a polar bear, but not about the small bug being crushed under its foot. While a penguin's brain may process information about both, it is not equally likely to store both experiences and the contexts in which they took place.

When discussing how information is stored, one becomes steeped in complex biological pathways. Such pathways involve proteins interacting with other proteins, proteins interacting with DNA, and the production of new proteins. In the neurobiological literature, much of the discussion around learning takes place at the level of synapses – synapses being the points where two neurons interact. The axon from neuron A attaches to a dendrite from neuron B and thus form a synapse. Synapses are the locus of neuronal communication in the brain. At an abstract level, learning is realised by information affecting neurons such that they progress through a sequence of stages: generation, stabalisation, consolidation, and maintenance [see @rudy_neurobiology_2014 for a digestible overview of each stage]. There are complicated mechanisms involved in each stage of learning, and not all the components are fully understood. Despite that, the literature provides a good sense of many important molecular events that are thought to be involved in the process of storing information in the brain so that it can be accessed in the future.

When some new bit of information has been processed by the brain, the physical embodiment in the brain is often described as the "memory trace" [@asok_molecular_2019; @robins_21st_2023; @semon_mneme_1921]. An influx of calcium (resulting from stimulation by an upstream neuron A) is necessary for the initial **generation** of the memory trace [@lynch_intracellular_1983; @cavazzini_ca2_2005]. The calcium interacts within the post-synaptic dendritic spine on neuron B and leads to actin being disassembled into smaller chunks [@lynch_ltp_2007]. Actin helps give structure to the spine. Breaking it down is a necessary step to create room to fit more receptors into the post-synaptic membrane of neuron B – the membrane acts as the skin between the cell and the watery world outside of the cell. Adding receptors to the membrane makes the spine on neuron B more sensitive to the upstream firing of neuron A, and thus more likely to itself fire an action potential in response. This rapid change in sensitivity (potentiation) is short lived. Without further processing, the potentiated state will revert back to baseline.

**Stabilisation** of the trace requires expanding and strengthening the actin network in the dendritic spine to solidify the modifications [@chen_changes_2007]. The spine head is enlarged, and the actin scaffolding is altered to make it less vulnerable to being broken down (via depolymerisation). Furthermore, proteins in the cell membrane called adhesion molecules help to couple the pre- and post-synaptic neurons, improving the effectiveness of neurotransmission [@huntley_cadherin_2002]. These first two stages, generation and stabalisation, are the pencil marks of memory. But these scratchings must be committed to ink to stand the test of time.

**Consolidation**, the stage where initial changes are solidified to stave off forgetting, involves the production of new proteins. In response to neuronal stimulation, a number of events take place in the postynaptic neuron. One response is the production of proteins which are able to enter the nucleus. Once in the nucleus, these proteins bind to DNA. In association with other molecular signaling, this activity leads to transcription of new molecules (mRNA) that will then be turned into proteins, such as receptors that can be inserted in the membrane. This genomic signaling ensures new proteins are continually minted, providing a sufficient pool of receptors and other elements needed to keep the spine in a potentiated (more sensitive) state. During the subsequent **maintenance** stage this supply of membrane receptors is produced and inserted into the membrane. and the typical dynamics of receptor removal, where receptors are removed from the membrane and recycled, is slowed. More excitatory receptors remain on the cell membrane which ensures heightened sensitivity to signals from presynaptic neurons both now and in the future.

## Memory research in animals and invertebrates

Many organisms have been poked, prodded and leered at during our efforts to understand the mechanisms of memory. Scrub jays, a bird sporting a bold blue coat and pointed black beak, have been recurring subjects in studies investigating spatial memory. This is because of their food caching expertise [@shettleworth_how_1982]. Sophisticated methods have been used to study changes in hippocampal volume in response to caching. The possibility of season-dependent changes in hippocampal neurogenesis in caching birds has also been explored [reviewed in @pravosudov_25the_2007].

Rodents have featured heavily in the experimental memory literature [@ghafarimoghadam_review_2022]. Recent advances in stimulation and imaging, specifically techniques like optogenetics [@goshen_optogenetic_2014] and two-photon microscopy [@kawakami_vivo_2015], have enabled us to study representation of different types of memory at levels ranging from individual synapses to neuronal ensembles. The last two decades saw a growing interest in episodic memory in rodents. This is conceptualised as the ability to represent the past and draw on specific encoded events in a manner akin to mental time travel [@tulving_episodic_2002; @eacott_mental_2007; @krause_episodic_2022]. Intricate tasks have been developed which enable rats to demonstrate memory for the context in which a stimulus had been previously presented, and to disentangled this from mere familiarity with the stimulus due to temporal proximity [@panoz-brown_rats_2016]. However, the existence of episodic-like memory in non-human animals remains controversial [@tulving_episodic_2005; @hoerl_thinking_2019]. The establishment of procedures for identifying and manipulating complex episodic memory, by optogenetic and other means, may help inform us on the type of mechanisms (synaptic or molecular) that underpin episodic memory in humans.

Research in birds and rodents has supplied us with decades of insights into behaviour and the brain structures involved in memory. But when seeking to understand the precise molecular and structural changes that underpin memory, the field first looked to invertebrates. The simplicity of invertebrate neural architecture allowed researchers to account for and track the entirety of a nervous system, and to observe the functional specificity of individual neurons. But simplicity is not the only benefit. By using simpler organisms, costs can be significantly reduced and environmental variables can be more easily controlled. Moreover, ethical concerns are relieved (albeit not completely removed) due to the reduced likelihood of finding meaningful sentience at this level.

While perhaps largely unknown outside of the sciences, *Aplysia* is a celebrity among the invertebrates for its contribution to the neurobiology of learning. *Aplysia* is a marine snail with a simple nervous system. The abdominal ganglion (collection of neurons) of *Aplysia* is home to the largest known neurons in nature [@moroz_aplysia_2011]. This makes it an ideal candidate for electrophysiology – the approach where neural activity is recorded by inserting electrodes into cells or in the space surrounding cells. In *Aplysia* , stimulation of the siphon (used for transporting water in the body) leads to a defensive retraction of the gill [@carew_classical_1981]. Repeated stimulation leads to a decrease in the intensity and length of the response, a simple form of non-associative memory called habituation[@pinsker_habituation_1970]. This simple form of learning is not something we typically care about. Yet, even this basic form of learning in *Aplysia* served as a useful platform for understanding the general principles of of learning described above, from generation through to consolidation. It is thought that these stages are conserved across organisms, and underpin more complex forms of learning that are of interest to humans.

*C. elegans* is another organism which towers above most invertebrates in terms of popularity. *C. elegans* gained prestige after it was the first organisms to have its connectome mapped [@white_structure_1986]. Understating the wiring of all 302 neurons in *C. elegans* allowed for a systems perspective of the nervous system. We could piece together the role of each neuron in helping the body to perform actions such as navigation, digestion, and defensive behaviours. Though the scale *C. elegans* nervous system is small compared to mammals, it revealed principles of neuronal organisation which persist across brains of all sizes. Principles such as reciprocal inhibition to facilitate movement, computing at the level of the cell for efficiency, and minimising the total length of neuronal wire [@sterling_principles_2015]. Although we may find comfort in distancing ourselves from so-called "lower organisms", nature is indifferent to our need for preeminence. Our brains may be bigger, but nature has equipped us with many of the same basic processes for learning, navigating, and operating in a complex world.

The study of invertebrates revealed that even complex behaviour can arise from a modest number of neural cells. Consider that C. elegans has less connections in its entire nervous system (\~7000) than a single mammalian pyramidal neuron [@cook_whole-animal_2019; @sterling_principles_2015; @megias_total_2001]. Yet, this bare-bones neuronal setup is sufficient for detecting a variety of chemical and olfactory cues, navigating the environment, escaping threats and detecting dynamic environmental signals such as temperature changes and social crowding. And as we climb the ladder of complexity from *C. elegans* to more complex invertebrates, the cognitive capabilities and potential for translational insights expands in turn.

### Planaria as a model organism {#sec-planaria-as-a-model-organism}

Planaria are a broad group of invertebrates which have become the centerpiece for a number of scientific investigations. Planaria are being used to investigate topics including regenerative biology [@karami_planarians_2015], toxicology [@li_effects_2008; @hagstrom_comparative_2019], radioprotective materials [@ermakov_planarians_2021] addiction [@raffa_planaria_2008], and the effect of zero gravity environments on morphology [@vista_ssep_mission_11_team_studying_2018]. Planaria have built niches across many ecological contexts and can be found across salt-water, fresh-water and land environments. Land dwelling planaria span up to half a meter long [@esser_land_1981], whereas freshwater planaria, which are more commonly used in behavioural research, are less than a centimeter in length [@vasquez-doorman_current_2022].

Planaria are bilaterians. They display bilateral symmetry across their left and right sides [@sluys_planarian_2018]. Planaria exhibit anterior-posterior polarity, such that their head can be distinguished from the tail in both structure and their behavioural repertoire. While the tail end of a planarian is rather uninteresting, the head has many intriguing features. Auricles are what give the head of many planarian species a triangular shape. Auricles are thought to support the detection of food and noxious chemicals in the immediate environment [@asano_rhodopsin-like_1998]. Eyespots, which are the most discernible feature of planaria, sit atop the dorsal surface of the head. These light-sensitive cell clusters enable planaria to detect light intensity and its direction [@shettigar_hierarchies_2017].

Of particular interest to neuroscientists, the planarian head harbors a bilobed brain which is needed to coordinate activity throughout the body[@inoue_planarian_2015]. This simple neural structure is of special evolutionary significance as planaria are thought to be the oldest organism to house an organised central nervous system, or what we might label as a true brain [@pagan_first_2014; @sarnat_brain_1985]. In real terms, a planarian brain lacks many features compared to the exuberance of the mammalian brain. But in relative terms, the brain-to-body mass ratio of planaria is similar to that of a rat [@best_transphyletic_1983].

The planarian brain resembles a horseshoe[@agata_structure_1998; @sarnat_brain_1985] and has been estimated to contain between twenty to thirty thousand neurons [@inoue_functional_2017, p.82]. The brain exhibits nine branches on each side which radiate out from the center. The lobes of the brain form thin nerve cords at their posterior end. These chords extend down the length of the body towards the tail, and together with the brain, comprise the central nervous system. The left and right nerve cords are connected by commissures which form a ladder-like structure [@sluys_planarian_2018]. At first glance, the nerve chords do not have the organisational complexity of the planarian brain. Judging purely from appearances, the nerve chords should not be capable of performing complex behaviour. However, it has been shown that even cells in the nerve chords can take on the behavioural profile of a head, indicating that the nerve chords are not simply axons innervating muscle [@le_planarian_2021].

Planarian neurons appear more similar in structure to those of vertebrates than of other invertebrates [@sarnat_brain_1985]. They feature spine-like protrusions on dendrites [@sarnat_brain_1985; @petralia_diversity_2016], and many neurons contain many dendritic branches but a singular axon. Zooming in further, planarian neurons contain a variety of synaptic vesicles, such as clear and dense-core variations, which resemble those seen in vertebrate neurons [@oosaki_observations_1965]. Most relevant to the research described in this project, planaria produce many of the same neurotransmitters and neuromodulators that we humans' possess. These include serotonin, dopamine, epinephrine, acetylcholine, GABA, glutamate and opioid peptides [@sarnat_brain_1985; @welsh_monoamine-containing_1970; @rawls_measurement_2006; for a comprehensive review of planarian neurochemistry see @buttarelli_neuropharmacology_2008].

The physiologist August Krogh posited that "You will find in the lower animals mechanisms and adaptations of exquisite beauty and the most surprising character" [-@krogh_progress_1929, p.203]. The conservation of neurochemistry in planaria is intriguing. But the primary feature making planaria worthy of Krogh's dictum is their regenerative capability. Planaria undergo a natural form of fission as part of their reproductive cycle. They tear themselves in half, with each half then regrowing all the necessary parts of its basic body plan to form a complete planarian again. A form of reproductive immortality. Regeneration is not completely novel in nature. Humans can regrow skin, and salamanders can regrow amputated limbs. But what sets planaria apart from the rest of the natural world is its ability to regrow its brain and central nervous system.

Planarian regeneration is facilitated by adult pluripotent neoblast cells which are found throughout the body [@neuhof_vertically-_2016]. After significant injury, these cells proliferate and undergo differentiation, providing the cell types needed to restore organs, membranes, and neural networks in the brain. This capability has drawn interest from medical researchers for more than a century [@morgan_experimental_1898; @child_patterns_1941; @reddien_cellular_2018]. By understanding the factors that control planarian regeneration, we may be able to artificially simulate these processes in humans to restore limbs or neural structures after injury. Business ventures have already been established in this area. Companies such as Morphoceuticals are looking to apply lessons learnt from planarian regeneration to rodents and, pending pre-clinical success, eventually humans.

### Review of planarian memory literature

In the 1900's, there was still some debate regarding whether invertebrates have the cognitive means needed to learn. A skeptical approach was evident from Donald Jensen in the 1970s who posited that "no invertebrate, no matter how complex is capable of showing 'true learning'" [quoted in @rilling_mystery_1996, p. 591]. This view established an artificial barrier separating organisms that suitably model human cognition from those that do not. Because invertebrates were overlooked, researchers tried to make progress on the neurobiology of memory using the complex nervous systems of rodents. After many years searching for the rodent engram (the collection of neurons underlying a specific learning event), this venture unearthed little of value [@mcconnell_manual_1967, pp. 2-3]. A group of psychologists including James McConnell in the 1970s were aware that little progress was being made in this endeavor. The group moved defiantly away from rodents and drifted towards invertebrates. Starting with much simpler organisms would allow researchers to progress past mere descriptions and arrive at actual understanding of the mechanisms of learning. Starting with rodents was akin to assembling a rocket ship for exploring space before figuring out commercial aviation within the earths atmosphere.

At first, McConnell and colleagues completed basic experiments showing that planaria could learn to associate a light with a shock [@mcconnell_effects_1959]. Compared to control subjects, trained planaria would exhibit more body contractions in response to light (a conditioned stimulus) and perform more changes of direction. But criticism arose over the lack of controls in these experiments [@travis_replicating_1981]. Later follow ups included blinding the experimenter and testing for confounding factors such as pseudoconditioning (the eliciting of the unconditioned response by other stimuli despite there being no relationship between them) and sensitisation (an increase in responding to a CS due to repeated presentation, rather than because of its association with a US). Contrary to the expectations of psychologists at the time, evidence for learning in invertebrates accrued study after study. It was eventually impossible to deny the capacity for stable associative memories to these (admittedly) primitive looking creatures. McConnell and others such as Eric Kandel established definitively that invertebrates are capable of learning, retaining, and acting on information.

The ability to form associative memories is an impressive feat given the bare-bones neural networks of the planarian brain. But why pursue planaria as a model organism? Why not shift all our resources towards other sophisticated invertebrates like honeybees and fruit flies? It was the pairing of this widely held capacity to learn with the rare ability for regeneration in planaria that sprouted one of the most peculiar branches of scientific endeavor to date. Investigating the retention of memory after decapitation and regeneration of the brain. Because planaria could form associations, researchers were able to ask the question: What happens if you teach a planarian, then cut it in half? Does the tail, which needs to regenerate its head and central nervous system, retain any of that prior learning? James McConnell, alongside Allan Jacobson and Daniel Kimble were the first scientists to pose and pursue the answer to this question [@mcconnell_effects_1959]. Across a range of different training procedures, McConnell and colleagues found that planarian tails indeed retain information, suggesting the intuition that memories must be stored in the brain may be mistaken, at least in some cases [@mcconnell_manual_1967]. Through some mechanism, memories are stored or backed up outside the brain and can be reinstantiated in the new brain during regeneration.

After the McConnell chronicles of the early 60s, invertebrate memory research saw its modern resurgence in the work of Shomrat and Levin [-@shomrat_automated_2013]. The authors published an important paper which utilised an automated training protocol for planaria. Planaria, like rodents, are hesitant to approach food in the center of a novel environment [@best_protopsychology_1963]. They will first explore the territory, and only then engage in consumption. As planaria become familiar with the environment through repeated trials, they begin to approach the food more quickly, demonstrating a form of recognition memory. The authors wanted to know whether this type of memory persists in the tails of trained planaria after they have been cut in half and had to regenerate a brain from scratch.

Planaria can detect and distinguish surface textures – an ability referred to as thigmotaxis. Shomrat and Levin's study therefore used smooth and rough surface textures for the conditioned stimuli. Over ten consecutive days, half of the planaria were fed on the target rough surface ("familiar" planaria) while the other half were only fed on a smooth surface ("naive" planaria). At the end of the training period, the familiar group took a significantly shorter amount of time to approach and consume the food in the rough environment. Both groups were then bisected into head and tail halves ("offspring") and left to regenerate for 10-14 days. The authors then looked at whether the tail offspring of familiar planaria retained familiarity of the rough environment and thus approached food relatively quickly compared to the naive tail offspring. The data suggested that regenerated fragments from familiar planaria did approach food more quickly, but this did not reach statistical significance. Moreover, after undergoing the same training procedure as the original planaria, the regenerated tail fragments from familiar planaria demonstrated a form of memory savings. The familiar tail sections became accustomed to the rough environment faster than tail sections of control (unfamiliar) planaria. This hints that some memory trace from prior training survived brain regeneration but required repetition of the training process for the memory savings to be expressed.

More recently, Samuel and colleagues [-@samuel_addiction-related_2021] corroborated this puzzling memory retention effect. The authors used sucrose to shift the surface preference of some planaria from their innate preference for a smooth surface to the sucrose-paired rough surface. After amputating the planaria and allowing time for head regeneration, it was observed that the tail halves retained the sucrose-paired rough preference, despite the newly regenerated brain never having been exposed to the rough surface. In contrast, the tail halves of control planaria – which were exposed to the rough surface but did not receive sucrose in this environment – showed the expected initial preference for the smooth surface.

One assumption implicit in planaria experiments to date is that, while a brain is not needed for storage of all memories, it is necessary to act upon the memories. For this reason, sufficient time is given for the brain to regenerate, enabling the planarian to swim and feed. However, a recent preprint by Shimojo and colleagues [-@shimojo_preservation_2022] goes against this assumption. They sought to test whether dissected tail halves of planaria can display a conditioned response (which was learnt before dissection) before the bran has a chance to regenerate. In this study, planaria were trained to associate a neutral weak UV light (conditioned stimulus) with an aversive shock (unconditioned stimulus). The shock typically causes planaria to twist their body – an unconditioned contortion response. After pairing the light with the shock, planaria will display a conditioned contortion response to the UV light alone. On the second and third day after dissection, well before the brain is thought to be reformed, the tail halves were exposed to the conditioned stimulus over a number of trials and their responses were recorded. The authors analysed the data using a deep neural network to classify behaviour. They found that most responses from the tail halves were similar to those produced by an electric shock rather than those produced by a neutral ultraviolet light. Ultimately, this suggests the tail halves retained the conditioned behaviour and were able to act on it, despite lacking a brain at the time.

Following a similar procedure, Rhodes and Vierick [-@rhodes_effects_2024] used classical conditioning to establish conditioned negative phototaxis (moving away from the conditioned light stimulus). Typically, planaria are strongly averse to blue light, mildly averse to green light, and are indifferent to red light [@paskin_planarian_2014]. Planaria were trained to associate a neutral red light with an aversive green light across 5 days. After conditioning, some planaria were bisected into head and tail halves. Three weeks later, the head and tails were tested for retention of the memory – negative phototaxis to red light. The results indicated that the bisected halves retained the conditioned memory just as well as intact planaria. Moreover, memory retention was not statistically different for regenerated heads and tails. This suggests planaria can retain a conditioned response for at least three weeks. What's more, the tail halves which lost their head and therefore had to regenerate a head and brain also retained and acted on this memory. One major issue with this study was that the number of planaria per group was very small, most containing just four to six subjects. Another key issue is it was not clear how much movement was necessary to support negative phototaxis on a given trial. While we must maintain skepticism for individual studies given their limitations, the number of findings showing successful retention of learning through regeneration provide strong support for the phenomena more generally.

Although classical conditioning procedures are common in the planarian literature, some experimenters have employed operant conditioning methods [@crawford_operant_1967; @chicas-mosier_new_2015; see @best_behavior_1963 for a review of early studies]. A simple learning procedure known as the Van Oye maze was one of the first forms of reinforcement learning in planaria [@van_oye_over_1920; @wells_training_1967; @raffa_analysis_2008]. In the typical setup, planaria are housed in a beaker and a fishing line with food attached is placed just below the water surface. Planaria can detect the presence of food and navigate towards it [@ash_chemical_1973; @miyamoto_chemotaxis_1985]. Planaria must navigate up the wall, across the surface and down the line to reach the food. This is a low probability behaviour, but some small percentage of planaria will reliably do this given enough trials. Control planaria undergo similar methods but without the food reward attached. At test, food is not placed on the rod, but is instead dissolved in the water beforehand. The dissolved food is a cue that food is available. Trained planaria are subsequently found in much greater numbers on the suspended line compared to control subjects. Across five experiments performed by Wells, an average of \~17 trained subjects were found on the line at test compared to an average of \~3 experimental subjects [reviewed in @corning_planarian_1970]. This procedure demonstrated that planaria can be trained using reinforcement learning.

Another operant conditioning study was conducted by Corning [-@corning_retention_1966] during the zenith of planaria fame. Corning wondered whether planaria can retain an operant conditioned behaviour after bisection and regeneration. Using a T-maze apparatus, planaria were trained to select their less preferred side across many trials. Positive reinforcement came in the form of being returned to the home arena for 10 minutes after making a correct choice. After incorrect choices they were taken to the start of the maze for another trial. Corning decided on a threshold for successful learning of nine out of ten consecutive correct choices across trials. Planaria that met this threshold were bisected.

After a two to three week regeneration period, the regenerates (both heads and tails) were given a baseline preference test and were subsequently conditioned to criterion. Corning found that the baseline of trained offspring differed significantly from the baseline of the original planaria, while untrained planaria offspring did not differ from the original subjects. This suggested operant conditioned behaviour can be retained outside of the planarian brain. Extending this, the regenerates of trained planaria could also be conditioned to threshold faster than regenerates of untrained planaria. A demonstration of direct memory recall and memory savings.

The expeiments above provide some evidence that planaria can learn conditioned responses. But when considering the persistence of this behaviour through bisection and regeneration, there is very limited evidence. Much of the research on operant conditioning in planaria dates back to the mid-tweintieth century. Although historical research still holds value, modern psychological science has raised questions regarding the reliability and replicability of past results.

Recent evidence suggests that the psychological literature broadly considered has oversold the existence and or size of many psychological phenomena [@open_science_collaboration_estimating_2015]. Much effort is being devoted towards identifying the types of decisions which lead to unreliable results appearing in journals [@simmons_false-positive_2011]. It is clear that exploiting researcher degrees of freedom – the design and analysis decisions taken which lead to untrustworthy results – has been a commonplace practice [@gopalakrishna_prevalence_2022]. While replication attempts may often focus on findings from the last two decades, we must also carry over this skepticism to research from the twentieth century. Especially in cases where there is only one or two reports of a given phenomenon. Given that, we should seek to establish reliable methods for inducing operant conditioned behaviours in planaria. Furthermore, we should withhold judgement on whether learned behaviours can persist through decapitation and regeneration in planaria until the phenomenon is replicated.

### Positive reinforcement of planarian behaviour

Investigators have used many different stimuli, both aversive and appetitive, in their efforts to condition planaria. One of the most common appetitive stimuli in the planarian literature is cocaine. Cocaine acts on dopamine receptors which are abundant in planaria, making it an ideal compound for positive reinforcement paradigms [@algeri_effects_1983; @buttarelli_neuropharmacology_2008]. Moreover, cocaine can be easily dissolved in water at low cost given the small quantity needed to reward planaria. The literature shows that cocaine induces strong effects on locomotion and atypical behaviours when given at high doses [@rawls_first_2010; @pagan_planarians_2013].

Cocaine is a stimulant which exerts it agonistic effects by blocking reuptake of dopamine through the dopamine transporters. In humans, this results in more dopamine activity in the synapse and therefore more neural activity in downstream neurons, particularly in the meso-limbic pathway connecting the ventral tegmental area to the nucleus accumbens [@nestler_molecular_2019]. This drug effect is linked to the rush or high that users experience, and is common to most if not all drugs of abuse. Cocaine also acts on serotonergic and noradrenergic transmission by blocking their respective transporters. The noradrenergic effects are thought to stimulate the sympathetic nervous system by blocking reuptake of norepinephrine and an decreased sympathetic nerve discharge, resulting in effects such as increased blood pressure and heart rate [@jacobsen_effects_1997; @freye_pharmacology_2009; @nestler_molecular_2019].

Amaning-Kwarteng et al. [-@amaning-kwarteng_relapse_2017] explored the establishment and extinction of a cocaine-reinforced texture preference, and the ability to reinstate this preference after extinction. They found that planaria can be conditioned using cocaine to shift their surface texture preference from smooth to rough, and that this preference can be extinguished (reverted back to the original preference) after repeated exposure without reinforcement. Subsequently, exposure to a bath of cocaine was enough to reinstate the conditioned preference when provided with both textures and given the choice of which surface it spends its time.

Building on prior work, Jawad, Hutchinson & Prados [-@mohammed_jawad_dissociation_2018] investigated addiction like learning and memory behaviour using sucrose. As is common in the planarian literature, the authors employed a conditioned place preference paradigm and measured conditioning, extinction, and tolerance. Conditioned place preference is commonly used to assess the motivational strength of drugs [@milton_persistence_2012]. A novel aspect of the work by Jawad and colleagues was the demonstration that appetitive learning in planarians requires dopaminergic activity, evidenced by a dopamine D1 antagonist blocking acquisition of a CPP but not interfering with tolerance. The use of planaria to understand addiction related phenomena such as tolerance and extinction has benefited from a recent resurgence. But its roots date back to at least the 1960's [@needleman_tolerance_1967].

Understanding the molecular changes and circuit dynamics underlying the establishment of addiction may illuminate a path towards chemical or behaviour therapies to change the brain and thus change behaviour. Currently, therapies focus on top down strategies. People are coached to recognise their thoughts and emotions related to drugs, and to manage them rather than simply act on them. However, if the chemistry and structural wiring of the brain change during the acquisition of an addiction, bottom up therapies may assist in reverting these maladaptive neuroadaptations [@chodkiewicz_conceptual_2023]. And although we may not be in a position to attempt extreme interventions in humans or other mammals, planaria enable us to pursue these questions and progress our understanding of how bottom up changes may reduce drug seeking behaviour.

## Unresolved questions

In the first half of the 20th century, there was doubt regarding whether invertebrates can learn. But as we look back nearly a century later, we have gathered ample evidence that planaria can form long-lasting memories [@wells_training_1967; @samuel_addiction-related_2021; @amaning-kwarteng_relapse_2017]. In addition, planaria serve as an especially useful organism due to their regenerative capabilities. Combined with conditioning procedures, there is now evidence that memory can be successfully retained in a planarian after having its head removed and thus having to regenerate a brain [@shomrat_automated_2013]. The persistence of associative memory despite loss of the brain is impressive. But even more incredible would be the persistence of learned behavioural responses.

An acquired texture preference is a valid form of learning. But it is far removed from the memories that concern us in our day to day lives. Instead, learning shaped by reward, requiring memory systems and dopamine circuits, better reflects the intentional learning we associate with intelligence in other humans. If complex memories formed by operant conditioning can persist in spite of total brain loss, this may have profound implications for the way we view memory storage and retrieval in humans. As discussed earlier, there is a lack of research investigating whether operantly conditioned behaviours can be maintained through decapitation and regeneration in planaria. This project aims to extend the phenomenon of memory retention through regeneration shown for classical conditioning to an operantly conditioned behaviour.

Shomrat and Levin [-@shomrat_automated_2013] observed that regenerates of planaria exposed to the rough surface did not initially differ in their performance compared to controls. However, it was clear that their performance on the task improved more rapidly than controls when they were exposed to the training procedure. What remains unknown is whether retraining is the only process that supports reinstantiation of the previously aquired memory. Could it be that other contextual cues, such as exposure to the reinforcer alone, are sufficient to bring back memories from the prior planaria?

The results of Shomrat and Levin [-@shomrat_automated_2013] suggested the memory was dormant but could be reawakened after exposure to the relevant training condition – a form of retention called "memory savings". Reactivation of a memory after prior failures is comparable to reinstatement effects in addiction research. After successfully training an animal to lever press for a reward such as cocaine, the behaviour can be extinguished by allowing the animal to repeatedly engage in the lever pressing behaviour without being rewarded [@de_wit_reinstatement_1981]. Eventually, the animal will stop performing the conditioned response when the lever is presented. However, if the animal is exposed to the reinforcer before being placed back in the operant chamber, the lever pressing behaviour will spontaneously return [@de_wit_reinstatement_1981].

With respect to both phenomena, the memory is either not accessible or is not acted upon, and requires exposure to the right stimulus to be reactivated. Although Extinction and reinstatement of drug seeking behaviour has been modeled in planaria [@amaning-kwarteng_relapse_2017], no experiments have explored whether a reinstatement phase can be used to reactivate memories which are dormant after decapitation and regeneration. The phenomena of savings demonstrates some memory trace is retained in the brainless tail half. Perhaps this trace can be reactivated without the need for retraining by instead exposing the planarian to the reinforcing stimulus. This project will therefore investigate whether memory stored outside the brain behaves like an extinguished memory in that mere exposure to the reinforcer is sufficient to reinstate the memory trace.

# Materials and Method

### Colony maintenance

The species' of planaria used in this project is not currently known. Due to restrictions on importing identified species such as *Schmidtea mediterranea* into New Zealand, local planaria were sourced from two local streams within Wellington, New Zealand (X stream and Y stream). Given the basic characteristics of the planaria (colour, head shape etc.) it is thought that there is a combination of Cura and Neppia species – both of which are commonly found in New Zealand waterways. We intend to perform genomic analysis at a later date to confirm the species identity. Planaria were housed in a 50 liter glass aquarium with internal filtering. The aquarium contained a natural ecological environment (rocks, snails, algae etc.). The tank water was maintained with -\*describe water treatment protocol\*- (referred to as "planaria water" hereafter). The room containing the aquarium was maintained at 23°C. The planaria were fed between one and three times a week, with meals consisting of \[describe food source\]. The colony was maintained on a 12-hour light/dark cycle with lights on at 9:30am till 9:30pm.

### Materials

**Experiment 1** (dose-response analysis): plastic petri dishes with a diameter of 5.5cm were used to assess motility. Petri dishes contained a final solution of 8ml, made up of planaria water for control subjects and cocaine hydrochloride (Sigma Aldrich, United States??-confirm) mixed with planaria water for experimental subjects. Planaria locomotion was captured using an OPPO A17 smart phone and the videos were imported into EthoVision (Noldus Information Technologies, Wageningen, the Netherlands) for motility tracking. Subjects were not visible at all times in the videos due to being occluded by a shadow or dish wall (\~10% of frames on average across all subjects). To resolve this issue, missing location data were interpolated using the interpolation feature in EthoVision. This imposes a direct line from the subjects last location to the next observed location to determine the distance traveled.

**Experiment 2** (Y-maze conditioning): three custom Y-mazes were used. The identical mazes were laser etched into 80x80mm plastic squares. Each maze contains a 27mm long runway, and two arms each being 25mm long. All compartments were 6mm wide and all walls were 6.5mm high. At the intersection between the runway and the arms, there is a small divot on the floor of the maze. This allows a plug 7mm in diameter to be inserted to trap liquid in the arms and enable controlled administration of a drug. The maze floor contained subtle lines as a result of the etching process. At the base of the runway there was a small externally powered white light (\~20 lux) which was fixed into the plastic. Light is an aversive stimulus which induces negative phototaxis and so should discourage planaria from resting at the start of the runway.

### Handling

Planaria were handled using different techniques for different circumstances. When removing planaria from their 12-well-plate, a filbert (medium length flat) paintbrush was preferred. However, when moving planaria between petri dishes and the y-maze, a fine artist's paintbrush was preferred. In other cases, such as when planaria would sit in the middle of the y-maze divot, a plastic transfer pipette with the tip cut off was used. Planaria were handled very gently throughout their lifespan. Rough handling was suspected to have caused a very high mortality rate during pilot experiments. This was likely due to puncturing the planaria by pointing the bristles at the body on approach. For the studies reported here, a different technique was adopted whereby the side of the bristles was used to gently scoop the planaria up the side of the wall until they attached their cilia to the brush.

## Experiment 1

Experiment 1 sought an appropriate dose of cocaine which was rewarding to planaria while not drastically altering their their locomotive behaviour. Cocaine has been frequently used as a rewarding compound to classically condition planaria and to investigate its toxicity [@amaning-kwarteng_relapse_2017; @raffa_description_2005; @tallarida_ethanol_2014; @hutchinson_persistent_2015; @palladini_pharmacological_1996]. Doses used for conditioning have ranged from 1μM [@hutchinson_persistent_2015] to 80μM [@raffa_cocaine_2005]. The species featured across conditioning experiments has varied from *Dugesia Tigrina* to *Dugesia gonocephala* and *Dugesia dorotocephala*.

Investigators have observed that some species are more amenable to conditioning procedures than others [@samuel_addiction-related_2021; @mueller_use_2002]. Differences in the behaviour and responses of planarian species have been observed in response to several types of stimuli [@cochet-escartin_scrunching_2015; @debold_differences_1965]. The species used throughout this project likely differs from those used elsewhere in the literature and may in fact be a species indigenous to New Zealand. For this reason, it is important to identify a suitable dose of cocaine which is rewarding but does not significantly alter motility.

60 planaria were used in this experiment. Planaria were assigned to one of six conditions formed from doses commonly used in the planarian conditioning literature. This included 0, 1, 5, 10, 20 and 100μM (*n* = 10 per condition). Subjects were run across twelve session. The first ten sessions each contained one subject from the 0 - 20μM conditions, whereas the last two session only contained subjects in the 100μM condition[^1]. Subjects were collected from the breeding tank on the day of data collection. Within each session, subjects were randomly allocated to their condition using a freely available [random number generator](https://stattrek.com/statistics/random-number-generator#table).

[^1]: initially only 0 - 20μM conditions were run. The 100μM condition was added after the initial data were analysed to ensure that the cocaine was having some effect on the planaria and was not inert.

Each dose-response session lasted 15 minutes. Prior to the first recording session of the day the drug concentrations were achieved by mixing cocaine (dissolved in distilled water) with planarian water to reach a final solution of 8ml. Each solution was mixed and allowed to sit for several minutes to ensure diffusion of the drug. Planaria were picked and assigned to a condition from a large pool prior to the session. A planarian was picked up and a random number sequence was followed to determine which condition it was assigned to. The recording began once all five subjects were in their respective dishes. After completing a single trial, planaria were rehoused in a large tank and were not used for any subsequent experiments in this manuscript.

@fig-figure2 provides an overview of the apparatus. Five petri dishes were positioned on a white acrylic sheet. Recording sessions took place under red light, with the light positioned 36cm above the dishes. The dishes were aligned in a 2x3 grid, with a gap left in the top middle position. The overhead light was centered here to minimise shadows cast over the dishes – this was important for digital tracking accuracy. Each drug concentration was rotated across the 5 grid positions between trials to control for any effects of lighting angle.

```{r, eval = TRUE}
#| label: fig-figure2
#| fig-cap: "Dose response apparatus"
#| include: true
#| fig-width: 12
#| fig-height: 7
#| fig-align: "center"
#Figure2
Figure2 <- readJPEG("Figures/Figure2.jpg") 
grid.raster(Figure2)
```

### Results and discussion

```{r}
#| label: Dose Response Descriptives
#| include: False

library(papaja)
library(rstatix)
library(afex)


# Source the script
source("Descriptives.R")
source("Graphs.R")
source("PvalueFormattingFunction.R")

```

@fig-boxplot depicts the distance moved by planarians across the six conditions. Prior to performing any statistics, the assumptions of normality and homogeneity of variances were tested. Levene's test for homogeneity of variances suggests there were equal variances across conditions (F = `r format(levene_dose_response$F, digits = 3)`, `r format_pvalue(levene_dose_response$P)`). The Shapiro-Wilk test results indicate that the data were not normally distributed (W = `r format(shapiro_test_dose_response$statistic, digits = 3)`,`r format_pvalue(shapiro_test_dose_response$p.value, digits = 3)`). Due to violation of the assumptions of ANOVA, a Kuskal test was performed. The results indicated that the effect of condition on distance is statistically significant (χ2 (`r kruskal_test$parameter`) = `r format(kruskal_test$statistic, digits = 3)`, `r format_pvalue(kruskal_test$p.value, digits = 3)`). An exploratory post-hoc Dunn's test was carried out to determine whether the 100μM group differed from other groups. The results indicated that the 100μM group differed significantly from several other groups: control (`r format_pvalue(dunn_test$P[11])`), 5μM (= `r format_pvalue(dunn_test$P[7])`), 10μM (`r format_pvalue(dunn_test$P[1])`), and 20μM (`r format_pvalue(dunn_test$P[4])`).

```{r, eval = TRUE}
#| label: fig-boxplot
#| fig-cap: "Plot of planarian motility by condition"
#| fig-width: 12
#| fig-height: 7
#| fig-align: center
#| warning: false
#| message: false

source("Plots/rainplot.R")
print(rainplot)

```

The results in @fig-boxplot convey the variability of planarian behaviour. All conditions had at least one subject which moved less than 30cm over the 15 minute recording, and all groups had at least two subjects that moved more than 140cm. Experimenter observations indicate that when placed in the recording dish, some planaria would move initially, and then come to rest within a few minutes at a spot on the wall. They would remain here without meaningful movement for the remainder of the recording. Although the 1μM group did not differ significantly from the control group, there is a curious grouping of planaria in the 1μM condition below 60cm. Consistent with this, Hutchinson et al., [-@hutchinson_persistent_2015] observed a significant decrease in motility during exposure to 1μM of cocaine but not to 10μM when compared to a control group. No potential explanation was offered for this unusual curvilinear pattern.

The results suggest a dose between 1μM and 20μM could be used for conditioning without negatively affecting planaria motility. It was also necessary to select a dose that is reinforcing for planaria. A range of cocaine doses have been used to successfully condition planaria in CPP paradigms. These procedures typically involve low doses such as 1μM [@hutchinson_persistent_2015], 5μM [@amaning-kwarteng_relapse_2017] or 10μM [@hutchinson_persistent_2015]. But in the case of CPP, exposure time per trial is relatively long, on the order of 15-20 minutes. Whereas In the operant conditioning paradigm proposed in Experiment 2, exposure time is just 3 minutes. To adjust for the smaller absorption window compared to CPP experiments, the larger 20μM concentration was chosen for experiment 2.

Alongside total motility, we were able to inspect how planaria motility changed over the recording interval. We observed that planaria moved more at the start of the session than towards the end, with a gradual decrease in the distance moved with each passing minute (see @fig-ridgeplot). An exploratory Welch's two sample t-test found a significant difference between the time traveled in the first minute (M = `r format(minute1_descriptives$mean, digits = 3)`, SD = `r format(minute1_descriptives$sd, digits = 3)`) compared to the 15th minute (M = `r format(minute15_descriptives$mean, digits = 3)`, SD = `r format(minute15_descriptives$sd, digits = 3)`), with subjects travelling significantly further during minute 15 (t(`r (time_binned_t_test_result$parameter)`) = `r format(time_binned_t_test_result$statistic, digits = 3)`, `r format_pvalue(time_binned_t_test_result$p.value)`). In future dose-response assessments, shorter recording sessions may suffice to assess dose-response curves.

```{r}
#| label: fig-ridgeplot
#| include: True
#| fig-cap: "Plot of planarian motility across recording interval"
#| apa-note: "Ridge plot of distanced moved by planaria during each minute interval. Each ridge shows the distribution of distances moved by all subjects during a given minute interval. Black bars indicate the mean distance moved for the whole sample."
#| fig-width: 12
#| fig-height: 7
#| fig-align: "center"

source("Plots/ridgeplot.R")

print(ridgeplot)

```

## Experiment 2

Next we sought to determine whether our species of planaria can learn and retain an operant conditioned response. Prior research demonstrates the capacity for learning in planaria by way of classical conditioning. It has been further shown that classically conditioned memories can be retained after decapitation and regeneration of the brain. But the capacity for complex memories shaped by operant conditioning to persist despite losing most of the central nervous system has not been definitively shown. As a first step towards testing retention of an operant conditioned memory through regeneration, we needed to first establish the capacity for operant learning in this species of planaria. This experiment was preregistered prior to data collection and can be found online at [Open Science Framework](https://osf.io/tq7u4/?view_only=9c794dd942fb4a54b6a986c0a893fe46) and at [PsycArchives](https://www.psycharchives.org/en/item/d6109ed1-9aab-467b-b981-e009be95f308).

Sixty planaria were used (treatment, *n* = 30; vehicle, *n* = 30). This experiment had four stages: baseline, conditioning, test, and reinstatement. During baseline and conditioning trials two planaria were run concurrently separate Y-mazes. Each maze was filled with 1.8ml of planaria water. Six planaria were used per run, wherein they completed either 6 trials (baseline) or 4 trials per day (conditioning) with an intertrial interval of approximately 15 minutes. At the start of each session six planaria were moved into holding petri dishes. At the start of a trial, two planaria were transferred to the middle of the maze runway using a paintbrush. A timer was stared once the planarian was placed in the runway. Planaria were given three minutes to enter one of the arms [^2]. Once a planarian had entered an arm, the plug was inserted to stop liquid moving between compartments, after which 0.5ml remained in each arm. A planarian was considered to have entered the arm when the plug could be safely inserted without touching the planarian.

[^2]: If the planarian had some part of their body in an arm, they would be given up to an extra minute to make their decision

When treatment subjects entered the active arm, 43.5μL of cocaine solution was pipetted near the center of the arm to achieve a 20μM concentration. If the inactive arm was selected, an identical volume of distilled water was pipetted near the center of the arm. After administration, the timer was restarted and three minutes were given for absorption. For control subjects, either arm resulted in plugging and then pipetting 43.5μL of distilled water into the arm. If a subject failed to enter an arm, the plug was inserted and 43.5μL of distilled water was pipetted near the subjects body while in the runway and then three minutes were given. The runway light was on throughout the duration of the trial. At the end of a trial, planaria were gently removed and placed back into their holding dish.

At test, six planaria were used per run. Three planaria were run concurrently in three separate Y-mazes. Planaria were given three minutes to enter an arm. Once a decision was made, the plug was inserted and planaria were left for approximately 60 seconds before being moved back to the holding dish. No additional liquid was added during test trials. The next group of three planaria would then begin their first test trial. The inter trial interval was approximately six minutes and thirty seconds. For the priming stage, the procedure was identical to the test stage, with the only additional step being drug exposure before the first trial. At the start of a run, the planaria were placed in a 20μM 8ml solution of cocaine diluted in planaria water for 10 minutes[^3]. At the end of the exposure interval, planaria were moved individually into a Y-maze to begin their first trial. Planaria were only exposed to cocaine before the first reinstatement trial, but not before subsequent trials.

[^3]: To optimise efficiency, the start time for each group of 3 was staggered so that subjects 1-3 had finished their first trial before the 10-minute period ended for subjects 4-6.

There were three exclusion criteria identified in the experiment preregistration. The exclusion criteria were: A) failing to complete at least four of the six baseline trials ; B) failing to complete at least two trials on consecutive conditioning days; C) failing to complete at least four of the last six trials. We attempted to replace all subjects excluded due to criterion A. However, due to time constraints, of the 18 that failed to meet this criterion, only 13 could be successfully replaced. Five subjects could not be replaced and so started conditioning despite having completed only two or three baseline trials. Thirteen subjects met criterion B or C and were excluded from the data analysis. There were no exclusion criteria set for test and reinstatement days. However, some subjects died in the waiting period, or demonstrated greatly impaired behaviour at test or reinstatement. Because of this, some subjects contributed data to baseline and endpoint, but not test or reinstatement, so the number of subjects in each condition differs across experimental phases.

```{r}
#| label: Experiment Two Results
#| include: False


# Source the scripts
source("ExperimentTwoAnalysis.R")
source("ExperimentTwoTimeAnalysis.R")


```

### Results and discussion

@fig-exp2decisions shows the average percentage of trials where subjects entered the active arm across four time points: baseline, endpoint, test, and reinstatement. A generalised linear mixed effects model with family set to binomial was fitted in R using the lme4 package [@bates_fitting_2015]. Subject ID was set as a random effect, with condition, time point and the interaction term as a fixed effects. Pairwise comparisons with a Bonferroni correction were carried out using the emmeans package in R [@lenth_emmeans_2024]. A Type III ANOVA was conducted using the using the car package [@fox_r_2019] to test identify whether there was a significant effects of condition, time, or an interaction.

The results did not find a significant effect of condition (χ2 (`r (exp2_decisions_model_output$Df[2])`) = `r format(exp2_decisions_model_output$Chisq[2], digits = 3)`, `r format_pvalue(exp2_decisions_model_output$"Pr(>Chisq)"[2])`). The results indicate a significant effect of time (χ2 (`r (exp2_decisions_model_output$Df[3])`) = `r format(exp2_decisions_model_output$Chisq[3], digits = 3)`, `r format_pvalue(exp2_decisions_model_output$"Pr(>Chisq)"[3])`) and a significant time\*condition interaction (χ2 (`r (exp2_decisions_model_output$Df[4])`) = `r format(exp2_decisions_model_output$Chisq[4], digits = 3)`, `r format_pvalue(exp2_decisions_model_output$"Pr(>Chisq)"[4])`).

Post-hoc pairwise comparisons were carried out using estimated marginal means with Bonferroni corrections applied to account for multiple comparisons. Comparisons looked at within group differences in response probability across the four phases and also at between group differences at each stage.The results indicated that there were two within-group differences for the control subjects: endpoint differed significantly from baseline (`r control_h_values_exp2_decisions %>% filter(contrast == "Baseline / Endpoint") %>% pull(apa_result)`), and test differed significantly from endpoint (`r control_h_values_exp2_decisions %>% filter(contrast == "Endpoint / Test") %>% pull(apa_result)`). There were three within-group differences for the treatment subjects: endpoint differed significantly from baseline (`r treatment_h_values_exp2_decisions %>% filter(contrast == "Baseline / Endpoint") %>% pull(apa_result)`), test differed significantly from baseline (`r treatment_h_values_exp2_decisions %>% filter(contrast == "Baseline / Test") %>% pull(apa_result)`), and reinstatement differed significantly from endpoint (`r treatment_h_values_exp2_decisions %>% filter(contrast == "Endpoint / Reinst") %>% pull(apa_result)`). A significant difference was found in the preference score between treatment and control subjects at test (`r between_group_h_final %>% filter(Time == "Test") %>% pull(apa_result)`).

\newpage

```{r}
#| label: fig-exp2decisions
#| include: True
#| fig-cap: "Mean Active Arm Preference Across Experimental Phases"
#| apa-note: "Changes in Y-maze active arm preference across experimental phases. The baseline phase included 6 trials, conditioning endpoint included the final 6 trials, and both test and reinstatement phases included 4 trials each. A) Control group showed significant differences in active arm selection between baseline and endpoint (p < .001), and between endpoint and test (p < .001). B) Treatment group demonstrated significant differences between baseline and endpoint (p < .01), baseline and test (p < .05), and endpoint and reinstatement (p < .05). C) Between-group comparison revealed significantly different active arm preferences during the test phase (p < .01). Error bars represent standard error of the mean. * = p <.05; ** = p <.01; *** = p <.001." 
#| fig-width: 12
#| fig-height: 14
#| fig-align: "center" 
#| dpi: 300

source("ExperimentTwoAnalysis.R")

print(combined_panel)

```

\newpage

Looking at the results in [@fig-exp2decisions]B, subjects in the treatment group demonstrate successful conditioning. These subjects were more likely to choose the active arm at the end of conditioning (endpoint) compared to baseline. Moreover, this preference was maintained for two weeks as evident by the heightened preference at test. Despite the increase in preference persisting for two weeks, when tested the next day during the reinstatement procedure the preference had returned to baseline levels.

[@fig-exp2decisions]A indicates the control group demonstrated a similar increase in preference for the active arm despite receiving no reinforcement. However, it contrast to the treatment group, the change in preference returned to baseline levels when tested two weeks after conditioning. This highlights the natural variability in planaria behaviour over time.

The between groups comparison seen in [@fig-exp2decisions]C shows a significant difference between groups at test. It may be that the preference stability shown by the treatment group is evidence of true learning as opposed to the natural variability of behaviour seen in the control group. This could explain why the treatment group effect remained even after two weeks, while the control group effect diminished back to baseline. But the reinforcement results constrain this rationale. If the conditioned behaviour was able to persist for several weeks, there is little reason to expect it would be extinguished rapidly but fail to show the expected effect of reinstatement. Overall, the results suggest planaria may have the capacity to learn this y-maze procedure. Future refinements may be needed to increase the rate of successful learning.

This experiment also considered the response time of planaria during each trial. We hypothesised that even if planaria cannot learn to make the correct decision, they may demonstrate increased motivation due to being aware that an award is available. This could be inferred from faster responding in the treatment group compared to the control group. The response time data across the four experimental phases are shown in [@fig-decision-time].

```{r}
#| label: fig-decision-time
#| include: True
#| fig-cap: "Mean Decision Latency Across Experimental Phases"
#| apa-note: "Mean time taken to make a response across phases and between conditions." 
#| fig-width: 16
#| fig-height: 14
#| fig-align: "center" 
#| dpi: 600

source("ExperimentTwoTimeAnalysis.R")

print(grouped_time_horizontal_boxplot)
```

\newpage

The decision latency data were analysed with a linear mixed-effects model using the lme4 package [@bates_fitting_2015]. The model included fixed effects of condition, time point, and an interaction term. Subject ID was set as a random effect to account for repeated measures. The decision time data were log-transformed. Type III ANOVA was conducted using the using the car package [@fox_r_2019] to test the significance of the fixed effects. Pairwise comparisons with a Bonferroni correction were carried out using the emmeans package in R [@lenth_emmeans_2024].

The decision latency results in @fig-decision-time show a significant effect of time (χ2 (`r (exp2_time_output$Df[3])`) = `r format(exp2_time_output$Chisq[3], digits = 3)`, `r format_pvalue(exp2_time_output$"Pr(>Chisq)"[3])`), but failed to show a significant effect of condition (χ2 (`r (exp2_time_output$Df[2])`) = `r format(exp2_time_output$Chisq[2], digits = 3)`, `r format_pvalue(exp2_time_output$"Pr(>Chisq)"[2])`). No time\*condition interaction effect was found (χ2 (`r (exp2_time_output$Df[4])`) = `r format(exp2_time_output$Chisq[4], digits = 3)`, `r format_pvalue(exp2_time_output$"Pr(>Chisq)"[4])`).

Post-hoc pairwise comparisons were carried out using estimated marginal means with Bonferroni corrections applied to account for multiple comparisons. Comparisons looked at within group differences in decision latency across the four phases and also looked at between group differences at each phase.The results indicated that there were two within-group differences for the control subjects: test differed significantly from baseline (`r control_results_exp2_time %>% filter(contrast == "Baseline / Test") %>% pull(apa_result)`) and test different significantly from endpoint (`r control_results_exp2_time %>% filter(contrast == "Endpoint / Test") %>% pull(apa_result)`). There were four within-group differences for the treatment subjects: test differed significantly from baseline (`r treatment_results_exp2_time %>% filter(contrast == "Baseline / Test") %>% pull(apa_result)`), reinstatement differed significantly from baseline (`r treatment_results_exp2_time %>% filter(contrast == "Baseline / Reinst") %>% pull(apa_result)`), test differed significantly from endpoint (`r treatment_results_exp2_time %>% filter(contrast == "Endpoint / Test") %>% pull(apa_result)`), and reinstatement differed significantly from endpoint (`r treatment_results_exp2_time %>% filter(contrast == "Endpoint / Reinst") %>% pull(apa_result)`).

The decision latency data do not support our hypothesis. Rather than decision latency decreasing over the experimental phases for the treatment group, there was no change for either treatment or control groups between baseline and endpoint. Moreover, during the follow up trials at test, both groups showed increased latency for decision making. It was observed that planaria became smaller across the duration of the experiment. This occurs as a result of being deprived of food for prolonged periods and is an adaptive response by planaria [@thommen_body_2019; @pascual-carreras_planarian_2020]. Although some research suggests planarian locomotor velocity is independent of body size [@raffa_quantitative_2001], this may only hold for between subjects comparisons during short experiments. In this case, when the size of planaria changes over time, it may be a sign of impaired health and low energy availability. While it may be reasonable to expect decreased decision latency across the first few conditioning days, as the experiment draws out, any effect of motivational may be diminished by the effect of starvation and energy depletion. Either there was no increase in motivation among planarians, or response latency may not be a suitable measure of planrian motivation.

\newpage

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::

## Graveyard

.Some common examples of invertebrates studied within psychology include flies, bees, worms, and octopi

Planarians are thought to have existed for approximately 300 million years (Vila-Farré & Rink, 2018).

This allowed fine-grained control of experimental variables to minimise variability across trials.
